---
title: "Predicting treatment response for people undergoing care for OUD"
author: "Nat Oliven"
date: last-modified
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predicting treatment response for people undergoing care for OUD}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    echo: false
    message: false
    warning: false
    error: false
    comment: ""
    dpi: 100
    fig-width: 6.5
    fig-height: 4.0
    R.options:
      digits: 3
editor: source
bibliography: [references.bib, packages.bib]
csl: apa.csl
---

```{r load-packages}
library(conflicted)
suppressPackageStartupMessages(library(tidymodels))
tidymodels_prefer()
suppressPackageStartupMessages(library(tidyverse))
library(public.ctn0094data)
library(gtsummary)
library(knitr)
conflicts_prefer(dplyr::filter)
options(dplyr.summarise.inform = FALSE)
```

```{r build-analysis}
#| include: false
source("building_analysis.R")
```

# Abstract

This study uses data from three NIDA Clinical Trials Network studies (CTN-0028, CTN-0030, CTN-0051) to build a predictive model of treatment response in people with opioid use disorder (OUD). We construct a pre-treatment feature matrix from behavioral, substance use, demographic, and clinical data in the 28 days before randomization, and evaluate whether these features predict relapse during the treatment period.

# Introduction

Opioid use disorder is associated with substantial mortality, morbidity, and economic burden in the United States. Medications for OUD (MOUD) — including methadone, buprenorphine-naloxone, and extended-release naltrexone — are effective treatments, yet individual response varies considerably. Matching patients to the treatment most likely to benefit them remains a central challenge in addiction medicine.

The CTN-0094 harmonized dataset combines three randomized clinical trials comparing MOUD regimens in opioid-dependent adults. By pooling data across studies, we have sufficient sample size to examine whether pre-treatment characteristics predict treatment response. The goal of this analysis is to build and evaluate a feature matrix capturing the 28 days prior to randomization, which will serve as input to predictive models of relapse outcomes.

# Methods

## Subjects

Data come from three NIDA Clinical Trials Network studies harmonized in the `public.ctn0094data` R package: CTN-0027 (methadone vs. buprenorphine-naloxone), CTN-0030 (buprenorphine-naloxone only), and CTN-0051 (extended-release naltrexone vs. buprenorphine-naloxone). Eligibility criteria for all three trials required a DSM-IV diagnosis of opioid dependence, age ≥ 18 years, and provision of informed consent. Only participants' first randomization is included; the CTN-0030 re-randomization step is excluded. The analysis sample includes `r nrow(analysis_base)` participants across `r n_distinct(analysis_base$project)` studies.

## Feature Engineering

All longitudinal features are derived from a 28-day pre-randomization window (days −28 inclusive to day 0 exclusive), providing exactly four of each weekday per participant. Substance use features are computed from the `all_drugs` dataset, which is more granular than the standalone `tlfb` table and covers multiple detection sources (TLFB self-report, urine drug screen, and urine drug screen with alcohol biomarker). Rare drug categories (<10 events in the window) and illicit categories used by fewer than 1% of participants are excluded before feature computation. Per-drug features include days of use, binary use indicator, and longest consecutive streak where clinically meaningful. Additional feature domains include demographics, self-reported behavior risk, addiction severity, psychiatric comorbidity, quality of life, pain, nicotine dependence, and withdrawal severity. Feature engineering is implemented in `building_analysis.R`.

## Statistical Methods

All models predict a binary relapse outcome from the pre-treatment feature matrix. Candidate models span interpretable baselines through high-capacity ensemble methods. Final model selection will be based on cross-validated area under the ROC curve (AUC-ROC).

### Logistic Regression

Standard binary logistic regression models the log-odds of relapse as a linear combination of predictors. Used as the interpretable baseline. Estimated via maximum likelihood; no regularization. Implemented in `tidymodels` using the `glm` engine.

### LASSO

L1-penalized logistic regression (Least Absolute Shrinkage and Selection Operator; Tibshirani, 1996) shrinks coefficients toward zero, performing automatic variable selection. Particularly useful given the high-dimensional feature matrix. The penalty parameter λ is selected via 10-fold cross-validation. Implemented via the `glmnet` engine in `tidymodels`.

### KNN

K-Nearest Neighbors classifies each observation by majority vote among its K nearest neighbors in feature space (Altman, 1992). Non-parametric; captures local non-linear structure without assuming a functional form. Features must be normalized to unit variance before distance computation. K is selected via cross-validation. Implemented via the `kknn` engine in `tidymodels`.

### CART

Classification and Regression Trees (Breiman et al., 1984) recursively partition the feature space by binary splits that minimize node impurity (Gini index). The resulting tree is interpretable as a set of if–then decision rules. Tree depth is controlled by cost-complexity pruning; the complexity parameter is selected via cross-validation. Implemented via the `rpart` engine in `tidymodels`.

### Random Forest

Random forests (Breiman, 2001) aggregate predictions from a large number of decorrelated decision trees, each trained on a bootstrap sample with a random subset of features at each split. The ensemble substantially reduces variance relative to a single tree. Number of trees, features per split, and minimum node size are tuned via cross-validation. Implemented via the `ranger` engine in `tidymodels`.

### Boosted Trees

Gradient boosted trees (Friedman, 2001; Chen & Guestrin, 2016) build an additive ensemble by sequentially fitting trees to the residuals of the current ensemble. Typically achieves high predictive accuracy but requires careful regularization (learning rate, tree depth, subsampling fraction) to prevent overfitting. Hyperparameters tuned via cross-validation. Implemented via the `xgboost` engine in `tidymodels`.

## Software

Analyses were conducted with `r stringr::word(R.Version()$version.string, 1, 3)`. The following R packages were used:

```{r tbl-software}
#| tbl-cap: "R packages used in this analysis."

tibble::tribble(
  ~Package,               ~Version,                                           ~Purpose,
  "tidyverse",            as.character(packageVersion("tidyverse")),           "Data manipulation and visualization (ggplot2, dplyr, tidyr, readr, purrr, stringr)",
  "tidymodels",           as.character(packageVersion("tidymodels")),          "Modeling framework (recipes, parsnip, rsample, yardstick, tune, workflows)",
  "gtsummary",            as.character(packageVersion("gtsummary")),           "Summary tables",
  "knitr",                as.character(packageVersion("knitr")),               "Document rendering and table formatting",
  "public.ctn0094data",   as.character(packageVersion("public.ctn0094data")),  "CTN-0094 harmonized clinical trial data",
  "conflicted",           as.character(packageVersion("conflicted")),          "Explicit conflict resolution for package namespaces"
) |>
  kable(align = "lll", col.names = c("Package", "Version", "Purpose"))
```

The following packages will be added at modeling time: `glmnet` (LASSO/ridge), `kknn` (KNN), `rpart` (CART), `ranger` (Random Forest), `xgboost` (Boosted Trees). [@R-base; @R-tidyverse; @tidyverse2019; @R-gtsummary]

# Results

## Participants

```{r tbl-participants}
#| tbl-cap: "Randomized participants by CTN study."

analysis_base |>
  count(project) |>
  mutate(project = as.character(project)) |>
  kable(col.names = c("Study", "N"), align = "lr")
```

A total of `r nrow(analysis_base)` participants across `r n_distinct(analysis_base$project)` studies were included after restricting to first randomizations (@tbl-participants).

## Participant Characteristics

```{r tbl-demographics}
#| tbl-cap: "Baseline participant characteristics by study."

analysis_tibble |>
  select(project, age, is_male, race, is_hispanic,
         education, is_living_stable) |>
  mutate(
    is_male         = factor(is_male,         levels = c(1, 0), labels = c("Yes", "No")),
    is_hispanic     = factor(is_hispanic,     levels = c(1, 0), labels = c("Yes", "No")),
    is_living_stable = factor(is_living_stable, levels = c(1, 0), labels = c("Yes", "No"))
  ) |>
  tbl_summary(
    by = project,
    include = c(age, is_male, race, is_hispanic, education, is_living_stable),
    label = list(
      age              ~ "Age (years)",
      is_male          ~ "Male",
      race             ~ "Race",
      is_hispanic      ~ "Hispanic ethnicity",
      education        ~ "Education",
      is_living_stable ~ "Stable housing"
    ),
    missing = "no"
  ) |>
  modify_header(label = "") |>
  bold_labels()
```

## Drug Use in the Pre-Treatment Window

```{r tbl-drug-prevalence}
#| tbl-cap: "Substance use categories observed in the 28 days before randomization (days −28 to −1). N users and percentage are relative to the analysis sample."

n_analysis <- nrow(analysis_base)

all_drugs_filtered |>
  semi_join(analysis_base, by = "who") |>
  distinct(who, what_grouped) |>
  group_by(what_grouped) |>
  summarize(n_users = n(), .groups = "drop") |>
  mutate(
    pct = round(n_users / n_analysis * 100, 1),
    what_grouped = as.character(what_grouped)
  ) |>
  arrange(desc(n_users)) |>
  kable(
    col.names = c("Substance category", "N users", "% of sample"),
    align = "lrr"
  )
```

## Predictor Distributions

```{r fig-withdrawal-severity}
#| fig-cap: "Pre-induction withdrawal severity by CTN study. Scores are derived from the COWS (CTN-0027/0030) or SOWS (CTN-0051) assessment completed before each participant's first MOUD dose (`withdrawal_pre_post`, what == 'pre'). 0 = None, 1 = Mild, 2 = Moderate, 3 = Severe. Participants with no pre-induction assessment recorded are excluded from this figure (n = `r sum(is.na(analysis_tibble$withdrawal_pre_score))` excluded)."
#| fig-width: 6.5
#| fig-height: 4.0

analysis_tibble |>
  filter(!is.na(withdrawal_pre_score)) |>
  mutate(
    severity = factor(
      withdrawal_pre_score,
      levels = 0:3,
      labels = c("None (0)", "Mild (1)", "Moderate (2)", "Severe (3)")
    ),
    project = as.character(project)
  ) |>
  count(project, severity) |>
  group_by(project) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = severity, y = pct, fill = project)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = "Pre-induction withdrawal severity",
    y = "Percentage of participants (%)",
    fill = "Study"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

## Feature Matrix

```{r feature-dimensions}
n_participants <- nrow(analysis_tibble)
n_features     <- ncol(analysis_tibble) - 2L   # subtract who and project
```

```{r tbl-dataset-sample}
#| tbl-cap: "First six rows of the analysis tibble (selected columns). The full matrix contains `r n_participants` participants and `r n_features` feature columns derived from pre-treatment behavioral and clinical data."

analysis_tibble |>
  select(
    who, project, age, is_male,
    heroin_days, pain_score,
    withdrawal_pre_score, has_major_dep
  ) |>
  slice_head(n = 6) |>
  kable(
    col.names = c(
      "ID", "Study", "Age", "Male",
      "Heroin days", "Pain score",
      "Withdrawal score", "Major depression"
    ),
    align = "llrrrrrr"
  )
```

The analysis tibble contains **`r n_participants` participants** and **`r n_features` feature columns** (excluding participant ID and study assignment). Feature domains span substance use history (per-drug days, binary use, longest streak), polydrug days, drug use breadth, demographics, addiction severity (ASI), nicotine dependence (Fagerström), pain (baseline scalar), psychiatric comorbidity, quality of life, housing stability, heroin denial (TLFB × self-report inconsistency), IV drug use risk behavior, withdrawal severity (pre-induction COWS/SOWS scalar), study treatment arm, site, prescribed medication adherence, and cross-domain interactions.

# Discussion

The feature matrix described here represents the pre-treatment behavioral and clinical profile for each randomized participant across the three CTN studies. The full feature set spans substance use, demographics, psychiatric comorbidity, pain, nicotine dependence, social risk behavior, IV drug use, withdrawal severity, medication adherence, and study site. These features will be used to train and evaluate predictive models of relapse, with the aim of informing treatment matching for OUD.

# References {.unnumbered}

```{r make-references}
#| include: false

knitr::write_bib(
  c(.packages(), "rUM"),
  "packages.bib"
)
```
