---
title: "Predicting treatment response for people undergoing care for OUD"
author: "Nat Oliven"
date: last-modified
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predicting treatment response for people undergoing care for OUD}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    echo: false
    message: false
    warning: false
    error: false
    comment: ""
    dpi: 100
    fig-width: 6.5
    fig-height: 4.0
    R.options:
      digits: 3
editor: source
bibliography: [references.bib, packages.bib]
csl: apa.csl
execute:
  cache: true
---

```{r load-packages}
library(conflicted)
suppressPackageStartupMessages(library(tidymodels))
tidymodels_prefer()
suppressPackageStartupMessages(library(tidyverse))
library(public.ctn0094data)
library(CTNote)
library(gtsummary)
library(knitr)
conflicts_prefer(dplyr::filter)
options(dplyr.summarise.inform = FALSE)
```

```{r build-analysis}
#| include: false
#| cache-extra: !expr digest::digest(readLines("building_analysis.R"))
# source() runs the entire script in place — every object it creates (analysis_tibble,
# analysis_base, all_drugs_filtered, etc.) lands in this document's R environment.
# It's not like Python's `from module import x` (selective). It's closer to
# exec(open("building_analysis.py").read()) — the whole file runs here, and
# everything it touches becomes available below.
# #| include: false means this chunk runs silently: no code, no output in the
# rendered document.
source("building_analysis.R")
```

```{r logistic-regression}
#| include: false
#| cache-extra: !expr digest::digest(readLines("logistic_regression.R"))
# Same pattern as build-analysis: runs logistic_regression.R in place, making
# all fitted model objects available to the Results section below.
source("logistic_regression.R")
```

```{r logistic-regression-wopv}
#| include: false
#| cache-extra: !expr digest::digest(readLines("logistic_regression_without_problem_vars.R"))
# Sources the cleaned model (problem variables excluded). The guard inside the
# script skips rebuilding analysis_tibble since logistic_regression.R already
# created it. All _wopv objects land in this document's R environment.
source("logistic_regression_without_problem_vars.R")
```

```{r logistic-regression-wopv-kfold}
#| include: false
#| cache-extra: !expr digest::digest(readLines("logistic_regression_wopv_kfold.R"))
source("logistic_regression_wopv_kfold.R")
```

# Abstract

This study uses data from three NIDA Clinical Trials Network studies (CTN-0027, CTN-0030, CTN-0051) to build a predictive model of treatment response in people with opioid use disorder (OUD). We construct a pre-treatment feature matrix from behavioral, substance use, demographic, and clinical data in the 28 days before randomization, and evaluate whether these features predict relapse during the treatment period.

# Introduction

Opioid use disorder is associated with substantial mortality, morbidity, and economic burden in the United States. Medications for OUD (MOUD) — including methadone, buprenorphine-naloxone, and extended-release naltrexone — are effective treatments, yet individual response varies considerably. Matching patients to the treatment most likely to benefit them remains a central challenge in addiction medicine.

The CTN-0094 harmonized dataset combines three randomized clinical trials comparing MOUD regimens in opioid-dependent adults. By pooling data across studies, we have sufficient sample size to examine whether pre-treatment characteristics predict treatment response. The goal of this analysis is to build and evaluate a feature matrix capturing the 28 days prior to randomization, which will serve as input to predictive models of relapse outcomes.

# Methods

## Subjects

Data come from three NIDA Clinical Trials Network studies harmonized in the `public.ctn0094data` R package: CTN-0027 (methadone vs. buprenorphine-naloxone), CTN-0030 (buprenorphine-naloxone only), and CTN-0051 (extended-release naltrexone vs. buprenorphine-naloxone). Eligibility criteria for all three trials required a DSM-IV diagnosis of opioid dependence, age ≥ 18 years, and provision of informed consent. Only participants' first randomization is included; the CTN-0030 re-randomization step is excluded. The analysis sample includes `r nrow(analysis_base)` participants across `r n_distinct(analysis_base$project)` studies.

## Feature Engineering

All longitudinal features are derived from a 28-day pre-randomization window (days −28 inclusive to day 0 exclusive), providing exactly four of each weekday per participant. Substance use features are computed from the `all_drugs` dataset, which is more granular than the standalone `tlfb` table and covers multiple detection sources (TLFB self-report, urine drug screen, and urine drug screen with alcohol biomarker). Rare drug categories (<10 events in the window) and illicit categories used by fewer than 1% of participants are excluded before feature computation. Per-drug features include days of use, binary use indicator, and longest consecutive streak where clinically meaningful. Additional feature domains include demographics, self-reported behavior risk, addiction severity, psychiatric comorbidity, quality of life, pain, nicotine dependence, and withdrawal severity. Feature engineering is implemented in `building_analysis.R`.

## Statistical Methods

All models predict a binary relapse outcome from the pre-treatment feature matrix. Candidate models span interpretable baselines through high-capacity ensemble methods. Final model selection will be based on cross-validated area under the ROC curve (AUC-ROC).

### Logistic Regression

Standard binary logistic regression models the log-odds of relapse as a linear combination of predictors. Used as the interpretable baseline. Estimated via maximum likelihood; no regularization. Implemented in `tidymodels` using the `glm` engine.

### LASSO

### KNN

### CART

### Random Forest

### Boosted Trees

## Software

Analyses were conducted with `r stringr::word(R.Version()$version.string, 1, 3)`. The following R packages were used:

```{r tbl-software}
#| tbl-cap: "R packages used in this analysis."

tibble::tribble(
  ~Package,               ~Version,                                           ~Purpose,
  "tidyverse",            as.character(packageVersion("tidyverse")),           "Data manipulation and visualization (ggplot2, dplyr, tidyr, readr, purrr, stringr)",
  "tidymodels",           as.character(packageVersion("tidymodels")),          "Modeling framework (recipes, parsnip, rsample, yardstick, tune, workflows)",
  "gtsummary",            as.character(packageVersion("gtsummary")),           "Summary tables",
  "knitr",                as.character(packageVersion("knitr")),               "Document rendering and table formatting",
  "public.ctn0094data",   as.character(packageVersion("public.ctn0094data")),  "CTN-0094 harmonized clinical trial data",
  "conflicted",           as.character(packageVersion("conflicted")),          "Explicit conflict resolution for package namespaces"
) |>
  kable(align = "lll", col.names = c("Package", "Version", "Purpose"))
```

The following packages will be added at modeling time: `glmnet` (LASSO/ridge), `kknn` (KNN), `rpart` (CART), `ranger` (Random Forest), `xgboost` (Boosted Trees). [@R-base; @R-tidyverse; @tidyverse2019; @R-gtsummary]

# Results

## Participants

```{r tbl-participants}
#| tbl-cap: "Randomized participants by CTN study."

analysis_base |>
  count(project) |>
  mutate(project = as.character(project)) |>
  kable(col.names = c("Study", "N"), align = "lr")
```

A total of `r nrow(analysis_base)` participants across `r n_distinct(analysis_base$project)` studies were included after restricting to first randomizations (@tbl-participants).

## Participant Characteristics

```{r tbl-demographics}
#| tbl-cap: "Baseline participant characteristics by study."

analysis_tibble |>
  select(project, age, is_male, race, is_hispanic,
         education, is_living_stable) |>
  mutate(
    is_male         = factor(is_male,         levels = c(1, 0), labels = c("Yes", "No")),
    is_hispanic     = factor(is_hispanic,     levels = c(1, 0), labels = c("Yes", "No")),
    is_living_stable = factor(is_living_stable, levels = c(1, 0), labels = c("Yes", "No"))
  ) |>
  tbl_summary(
    by = project,
    include = c(age, is_male, race, is_hispanic, education, is_living_stable),
    label = list(
      age              ~ "Age (years)",
      is_male          ~ "Male",
      race             ~ "Race",
      is_hispanic      ~ "Hispanic ethnicity",
      education        ~ "Education",
      is_living_stable ~ "Stable housing"
    ),
    missing = "no"
  ) |>
  modify_header(label = "") |>
  bold_labels()
```

## Drug Use in the Pre-Treatment Window

```{r tbl-drug-prevalence}
#| tbl-cap: "Substance use categories observed in the 28 days before randomization (days −28 to −1). N users and percentage are relative to the analysis sample."

n_analysis <- nrow(analysis_base)

all_drugs_filtered |>
  semi_join(analysis_base, by = "who") |>
  distinct(who, what_grouped) |>
  group_by(what_grouped) |>
  summarize(n_users = n(), .groups = "drop") |>
  mutate(
    pct = round(n_users / n_analysis * 100, 1),
    what_grouped = as.character(what_grouped)
  ) |>
  arrange(desc(n_users)) |>
  kable(
    col.names = c("Substance category", "N users", "% of sample"),
    align = "lrr"
  )
```

## Predictor Distributions

```{r fig-withdrawal-severity}
#| fig-cap: "Pre-induction withdrawal severity by CTN study. Scores are derived from the COWS (CTN-0027/0030) or SOWS (CTN-0051) assessment completed before each participant's first MOUD dose (`withdrawal_pre_post`, what == 'pre'). 0 = None, 1 = Mild, 2 = Moderate, 3 = Severe. Participants with no pre-induction assessment recorded are excluded from this figure (n = `r sum(is.na(analysis_tibble$withdrawal_pre_score))` excluded)."
#| fig-width: 6.5
#| fig-height: 4.0

analysis_tibble |>
  filter(!is.na(withdrawal_pre_score)) |>
  mutate(
    severity = factor(
      withdrawal_pre_score,
      levels = 0:3,
      labels = c("None (0)", "Mild (1)", "Moderate (2)", "Severe (3)")
    ),
    project = as.character(project)
  ) |>
  count(project, severity) |>
  group_by(project) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = severity, y = pct, fill = project)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = "Pre-induction withdrawal severity",
    y = "Percentage of participants (%)",
    fill = "Study"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

## Feature Matrix

```{r feature-dimensions}
n_participants <- nrow(analysis_tibble)
n_features     <- ncol(analysis_tibble) - 2L   # subtract who and project
```

```{r tbl-dataset-sample}
#| tbl-cap: "First six rows of the analysis tibble (selected columns). The full matrix contains `r n_participants` participants and `r n_features` feature columns derived from pre-treatment behavioral and clinical data."

analysis_tibble |>
  select(
    who, project, age, is_male,
    heroin_days, pain_score,
    withdrawal_pre_score, has_major_dep
  ) |>
  slice_head(n = 6) |>
  kable(
    col.names = c(
      "ID", "Study", "Age", "Male",
      "Heroin days", "Pain score",
      "Withdrawal score", "Major depression"
    ),
    align = "llrrrrrr"
  )
```

The analysis tibble contains **`r n_participants` participants** and **`r n_features` feature columns** (excluding participant ID and study assignment). Feature domains span substance use history (per-drug days, binary use, longest streak), polydrug days, drug use breadth, demographics, addiction severity (ASI), nicotine dependence (Fagerström), pain (baseline scalar), psychiatric comorbidity, quality of life, housing stability, heroin denial (TLFB × self-report inconsistency), IV drug use risk behavior, withdrawal severity (pre-induction COWS/SOWS scalar), study treatment arm, site, prescribed medication adherence, and cross-domain interactions.

## Logistic Regression

```{r tbl-lr-or}
#| tbl-cap: "Logistic regression odds ratios (p < 0.05, sorted by effect size). Wald 95% confidence intervals."

lr_or_table |>
  kable(
    digits    = 3,
    col.names = c("Predictor", "Odds Ratio", "95% CI Low", "95% CI High", "p-value")
  )
```

```{r tbl-lr-metrics}
#| tbl-cap: "Logistic regression model performance. Training and test sets from a stratified 75/25 split."

bind_cols(
  lr_metrics      |> rename(Train = .estimate),
  lr_test_metrics |> select(Test = .estimate)
) |>
  mutate(.metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) |>
  kable(
    digits    = 3,
    col.names = c("Metric", "Train", "Test")
  )
```

### Exclusion of Structurally Sparse Predictors

```{r wopv-ci-check}
#| include: false
wopv_ci_check <- lr_fit_wopv |>
  tidy(exponentiate = TRUE) |>
  mutate(
    conf.low    = exp(log(estimate) - 1.96 * std.error),
    conf.high   = exp(log(estimate) + 1.96 * std.error),
    log10_width = log10(conf.high / conf.low)
  ) |>
  filter(!is.na(estimate), term != "(Intercept)")

wopv_top3_ci <- wopv_ci_check |> slice_max(log10_width, n = 3)
```

A second logistic regression was fitted after excluding predictors identified as problematic on substantive or numerical grounds. Five predictors were excluded for leakage: `took_own_study_drug`, `took_other_study_drug`, `methadone_binary`, `buprenorphine_binary`, and `suboxone_binary` encode treatment arm assignment directly or by proxy, introducing post-randomization information into the pre-treatment predictor set.

The remaining exclusions addressed complete separation. Inspection of the baseline coefficient table revealed predictors with Wald confidence intervals spanning 25–90 orders of magnitude — the diagnostic signature of a non-existent maximum likelihood estimate. When a predictor nearly perfectly partitions the binary outcome, the log-likelihood has no finite maximum and the iteratively reweighted least squares optimizer terminates at an arbitrarily large value; the resulting point estimate is numerically meaningless regardless of its apparent plausibility. All such predictors were sparse drug-use indicators (analgesic, antiemetic, and muscle relaxant use and their binary counterparts) with very few observed events in the 28-day pre-treatment window.

Exclusion proceeded over three iterative rounds. In each round, all predictors with confidence intervals spanning more than approximately 10 orders of magnitude were removed, the model was refitted, and the coefficient table was re-examined. Predictors appearing in subsequent rounds arose exclusively from correlated pairs (Pearson *r* > 0.90 at the `step_corr` threshold): removing the first member of a pair permitted the second to pass the pairwise correlation filter, where it exhibited the same separation pattern. This sequential structure confirms that the final stable configuration reflects genuine structural sparsity rather than an incidental local solution — every excluded variable was either directly sparse or a downstream consequence of direct sparsity.

Following three rounds of exclusion totaling 15 predictors (beyond `treatment_arm`, which is excluded from both models), all `r nrow(wopv_ci_check)` estimable predictors had well-bounded Wald confidence intervals. The three widest spanned `r round(wopv_top3_ci$log10_width[1], 1)`, `r round(wopv_top3_ci$log10_width[2], 1)`, and `r round(wopv_top3_ci$log10_width[3], 1)` orders of magnitude (`r wopv_top3_ci$term[1]`, `r wopv_top3_ci$term[2]`, and `r wopv_top3_ci$term[3]`, respectively); all remaining intervals were narrower. Seven additional terms had missing coefficient estimates due to exact multicollinearity in the encoded design matrix (rank deficiency); these reflect structural properties of the one-hot and polynomial encodings and are not separation.

```{r tbl-lr-wopv-metrics}
#| tbl-cap: "Model performance after exclusion of structurally sparse predictors. Training and test sets from the same stratified 75/25 split (seed 12345) as the baseline logistic regression, enabling direct comparison."

bind_cols(
  lr_metrics_wopv      |> rename(Train = .estimate),
  lr_test_metrics_wopv |> select(Test = .estimate)
) |>
  mutate(.metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) |>
  kable(
    digits    = 3,
    col.names = c("Metric", "Train", "Test")
  )
```

### Cross-Validation

Ten-fold cross-validation was applied to the cleaned model on the training set to obtain a more stable performance estimate. Each fold held out approximately 10% of training observations (~187 participants); every training observation was held out exactly once. Cross-validated metrics are reported as means across the 10 folds (@tbl-lr-cv-metrics). Fold-to-fold variance in AUC was small, reflecting consistent model behavior across training subsets rather than sensitivity to any particular split.

```{r tbl-lr-cv-metrics}
#| tbl-cap: "Ten-fold cross-validated performance of the logistic regression model (without problem variables). Mean and standard error across folds."

rs_results |>
  collect_metrics() |>
  mutate(.metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) |>
  select(.metric, mean, std_err) |>
  kable(
    digits    = 3,
    col.names = c("Metric", "Mean", "SE")
  )
```

```{r fig-lr-cv-roc}
#| fig-cap: "ROC curves from 10-fold cross-validation. Gray lines show individual fold curves; black line shows the pooled overall curve."
#| fig-width: 5
#| fig-height: 5

fold_curves   <- rs_results |>
  collect_predictions() |>
  group_by(id) |>
  roc_curve(truth = outcome, .pred_0, event_level = "first")

overall_curve <- rs_results |>
  collect_predictions() |>
  roc_curve(truth = outcome, .pred_0, event_level = "first")

ggplot() +
  geom_path(
    data = fold_curves,
    aes(x = 1 - specificity, y = sensitivity, group = id),
    color = "gray70", linewidth = 0.5, alpha = 0.7
  ) +
  geom_path(
    data = overall_curve,
    aes(x = 1 - specificity, y = sensitivity),
    color = "black", linewidth = 1.2
  ) +
  geom_abline(lty = 3, color = "gray40") +
  coord_equal() +
  labs(x = "1 - Specificity", y = "Sensitivity") +
  theme_minimal(base_size = 11)
```

# Discussion

The feature matrix described here represents the pre-treatment behavioral and clinical profile for each randomized participant across the three CTN studies. The full feature set spans substance use, demographics, psychiatric comorbidity, pain, nicotine dependence, social risk behavior, IV drug use, withdrawal severity, medication adherence, and study site. These features will be used to train and evaluate predictive models of relapse, with the aim of informing treatment matching for OUD.

The preprocessing recipe handles two distinct classes of missingness separately: (1) structural MNAR missingness, where entire studies did not collect certain variables (e.g., substance use disorder diagnosis flags, homeless status), addressed by creating an explicit is-missing indicator column via `step_indicate_na()` before imputing, so the model can learn from the missingness pattern itself; and (2) random MCAR missingness, affecting a small number of participants across variables such as pain score and withdrawal severity, addressed by median or mode imputation with no indicator. Additional preprocessing steps include `step_ordinalscore()` for ordered factors with defensible equal spacing, polynomial contrast encoding for cigarettes-per-day, and `step_corr()` to remove predictors with pairwise correlation above 0.90, which eliminates redundant pairs such as the two withdrawal severity columns derived from the same single observation.

The baseline logistic regression achieved a ROC AUC of `r round(lr_metrics$.estimate[lr_metrics$.metric == "roc_auc"], 3)` on training and `r round(lr_test_metrics$.estimate[lr_test_metrics$.metric == "roc_auc"], 3)` on test — a mild to moderate drop consistent with overfitting expected from fitting an unregularized model to 100+ predictors, but still reflecting meaningful discriminative ability above chance. The large estimated associations for study membership (CTN-0030: OR ≈ 42, CTN-0027: OR ≈ 4 in the cleaned model) likely reflect between-study heterogeneity in outcome ascertainment or follow-up design rather than pre-treatment clinical effects, consistent with the challenge of pooling trials with distinct protocols.

# References {.unnumbered}

```{r make-references}
#| include: false

knitr::write_bib(
  c(.packages(), "rUM"),
  "packages.bib"
)
```
