---
title: "Predicting treatment response for people undergoing care for OUD"
author: "Nat Oliven"
date: last-modified
format:
  html:
    embed-resources: true
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predicting treatment response for people undergoing care for OUD}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    echo: false
    message: false
    warning: false
    error: false
    comment: ""
    dpi: 100
    fig-width: 6.5
    fig-height: 4.0
    R.options:
      digits: 3
editor: source
bibliography: [references.bib, packages.bib]
csl: apa.csl
---

```{r load-packages}
library(conflicted)
suppressPackageStartupMessages(library(tidymodels))
tidymodels_prefer()
suppressPackageStartupMessages(library(tidyverse))
library(public.ctn0094data)
library(CTNote)
library(gtsummary)
library(knitr)
conflicts_prefer(dplyr::filter)
options(dplyr.summarise.inform = FALSE)
```

```{r build-analysis}
#| include: false
# source() runs the entire script in place — every object it creates (analysis_tibble,
# analysis_base, all_drugs_filtered, etc.) lands in this document's R environment.
# It's not like Python's `from module import x` (selective). It's closer to
# exec(open("building_analysis.py").read()) — the whole file runs here, and
# everything it touches becomes available below.
# #| include: false means this chunk runs silently: no code, no output in the
# rendered document.
source("building_analysis.R")
```

```{r logistic-regression}
#| include: false
# Same pattern as build-analysis: runs logistic_regression.R in place, making
# all fitted model objects available to the Results section below.
source("logistic_regression.R")
```

```{r logistic-regression-wopv}
#| include: false
#| cache-extra: !expr digest::digest(readLines("logistic_regression_without_problem_vars.R"))
# Sources the cleaned model (problem variables excluded). The guard inside the
# script skips rebuilding analysis_tibble since logistic_regression.R already
# created it. All _wopv objects land in this document's R environment.
source("logistic_regression_without_problem_vars.R")
```

```{r logistic-regression-wopv-kfold}
#| include: false
#| cache: true
#| cache-extra: !expr digest::digest(readLines("logistic_regression_wopv_kfold.R"))
source("logistic_regression_wopv_kfold.R")
```

```{r knn}
#| include: false
#| cache: true
#| cache-extra: !expr digest::digest(readLines("knn.R"))
source("knn.R")
```

```{r knn-save-metrics}
#| include: false
# knn_wopv.R overwrites kknn_tune / favorite / relapse_pred_knn.
# Read scalar AUCs from the named objects knn.R exported before that overwrite. # added 6/2
knn_train_auc <- knn_train_metrics$.estimate[knn_train_metrics$.metric == "roc_auc"]
knn_cv_auc    <- knn_cv_metrics$mean[knn_cv_metrics$.metric == "roc_auc"]
```

```{r knn-wopv}
#| include: false
#| cache: true
#| cache-extra: !expr digest::digest(readLines("knn_wopv.R"))
source("knn_wopv.R")
```

```{r logistic-enet}
#| include: false
#| cache: true
#| cache-extra: !expr digest::digest(readLines("logistic_enet.R"))
source("logistic_enet.R")
```

```{r cart}
#| include: false
#| cache: true
#| cache-extra: !expr digest::digest(readLines("cart.R"))
source("cart.R")
```

```{r precompute-captions}
#| include: false
# Pre-compute captions that contain inline R calculations (commas inside
# paste0 break !expr YAML parsing when placed directly in chunk options). # added 6/2
cap_tbl_demographics <- paste0(
  "Baseline characteristics of the ",
  scales::comma(nrow(analysis_tibble)),
  " care-seeking adults with opioid use disorder in CTN-0027, CTN-0030",
  " and CTN-0051 (2006-2016), by study."
)
cap_tbl_drug <- paste0(
  "Substance-use categories observed in the 28 days before randomization",
  " (days -28 to -1) among the ",
  scales::comma(nrow(analysis_base)),
  " CTN-0027/0030/0051 participants (2006-2016).",
  " N users and percentage are relative to the analysis sample."
)
cap_fig_withdrawal <- paste0(
  "Pre-induction withdrawal severity among care-seeking adults with opioid",
  " use disorder in CTN-0027, CTN-0030 and CTN-0051 (2006-2016).",
  " Scores derive from COWS (CTN-0027/0030) or SOWS (CTN-0051) before",
  " each participant's first MOUD dose.",
  " 0 = None, 1 = Mild, 2 = Moderate, 3 = Severe.",
  " Participants with no pre-induction assessment are excluded (n = ",
  sum(is.na(analysis_tibble$withdrawal_pre_score)), " excluded)."
)
# cap_all_metrics retired — tbl-all-metrics now uses a static inline caption
# Read from named objects in knn_wopv.R (kknn_tune/favorite now = wopv versions). # added 6/2
knn_wopv_train_auc <- knn_wopv_train_metrics$.estimate[knn_wopv_train_metrics$.metric == "roc_auc"]
knn_wopv_cv_auc    <- knn_wopv_cv_metrics$mean[knn_wopv_cv_metrics$.metric == "roc_auc"]

cap_vip_glm    <- "Logistic-regression (glm engine) variable importance for predicting relapse among CTN-0027/0030/0051 participants (2006-2016), ranked by absolute z-statistic; problem variables excluded."
cap_vip_glmnet <- paste0(
  "LASSO (glmnet engine) variable importance: the ", nrow(lasso_coef),
  " predictors with non-zero coefficients at the selected penalty",
  " for relapse prediction among CTN-0027/0030/0051 participants (2006-2016)."
)
cap_vip_knn    <- "K-nearest-neighbors (kknn engine) permutation variable importance for relapse prediction among CTN-0027/0030/0051 participants (2006-2016): mean decrease in test ROC AUC when each predictor is shuffled (10 permutations)."
cap_vip_cart   <- "CART (rpart engine) variable importance for relapse prediction among CTN-0027/0030/0051 participants (2006-2016): native impurity-based importance, measured as the mean decrease in node impurity weighted by the number of training samples reaching each split."
```

# Abstract

This study uses data from three NIDA Clinical Trials Network studies (CTN-0027, CTN-0030, CTN-0051) to build a predictive model of treatment response in people with opioid use disorder (OUD). We construct a pre-treatment feature matrix from behavioral, substance use, demographic, and clinical data in the 28 days before randomization, and evaluate whether these features predict relapse during the treatment period.

# Introduction

Opioid use disorder is associated with substantial mortality, morbidity, and economic burden in the United States. Medications for OUD (MOUD) — including methadone, buprenorphine-naloxone, and extended-release naltrexone — are effective treatments, yet individual response varies considerably. Matching patients to the treatment most likely to benefit them remains a central challenge in addiction medicine.

The CTN-0094 harmonized dataset combines three randomized clinical trials comparing MOUD regimens in opioid-dependent adults [@ctn0027; @ctn0030; @ctn0051]. By pooling data across studies, we have sufficient sample size to examine whether pre-treatment characteristics predict treatment response. The goal of this analysis is to build and evaluate a feature matrix capturing the 28 days prior to randomization, which will serve as input to predictive models of relapse outcomes.

# Methods

## Subjects

Data come from three NIDA Clinical Trials Network studies harmonized in the `public.ctn0094data` R package: CTN-0027 (methadone vs. buprenorphine-naloxone), CTN-0030 (buprenorphine-naloxone only), and CTN-0051 (extended-release naltrexone vs. buprenorphine-naloxone). Eligibility criteria for all three trials required a DSM-IV diagnosis of opioid dependence, age ≥ 18 years, and provision of informed consent. Only participants' first randomization is included; the CTN-0030 re-randomization step is excluded. The analysis sample includes `r nrow(analysis_base) |> scales::comma()` care-seeking adults with opioid use disorder, drawn from `r n_distinct(analysis_base$project) |> xfun::n2w()` harmonized NIDA Clinical Trials Network studies (CTN-0027, CTN-0030 and CTN-0051) conducted between 2006 and 2016.

## Feature Engineering

All longitudinal features are derived from a 28-day pre-randomization window (days −28 inclusive to day 0 exclusive), providing exactly four of each weekday per participant. Substance use features are computed from the `all_drugs` dataset, which is more granular than the standalone `tlfb` table and covers multiple detection sources (TLFB self-report, urine drug screen, and urine drug screen with alcohol biomarker). Rare drug categories (<10 events in the window) and illicit categories used by fewer than 1% of participants are excluded before feature computation. Per-drug features include days of use, binary use indicator, and longest consecutive streak where clinically meaningful. Additional feature domains include demographics, self-reported behavior risk, addiction severity, psychiatric comorbidity, quality of life, pain, nicotine dependence, and withdrawal severity. Feature engineering is implemented in `building_analysis.R`.

## Statistical Methods

All models predict a binary relapse outcome from the pre-treatment feature matrix. Candidate models span interpretable baselines through high-capacity ensemble methods. Final model selection will be based on cross-validated area under the ROC curve (AUC-ROC).

All models share a single `recipes` preprocessing pipeline applied inside each `workflow`, so that resampling re-estimates every data-dependent step on the analysis fold only (no leakage). The recipe gives `who` an ID role; removes treatment-arm leakage variables (`treatment_arm` and its medication-type proxies) and zero-variance columns; handles missingness in two tiers — structural study-level missingness gets an explicit `step_indicate_na()` indicator before median/mode imputation, while sporadic item-level missingness is median/mode imputed without an indicator; scores ordered factors with `step_ordinalscore()` where equal spacing is defensible; standardizes all numeric predictors with `step_normalize()`; one-hot encodes unordered factors and uses polynomial contrasts for cigarettes-per-day; and finally drops predictors correlated above |r| = 0.90 with `step_corr()`.

::: {.callout-important title="NOTE TO SELF — REMOVE BEFORE SUBMISSION"}

**Metric-priority reasoning (settle before writing the Discussion):**

- **Model selection prioritizes AUC-ROC** — threshold-free, and the convention adopted here. AUC is *symmetric* in sensitivity and specificity, so it encodes no operating preference, and a sens/spec priority does **not** change which model is selected.

- **Specificity is the operating-point priority in this design.** With relapse coded as the positive class, specificity = correctly identifying the participants who will *not* relapse — i.e., **identifying people whose pre-treatment profile is genuinely promising for treatment**. The costly error to minimize is the false positive (flagging a true non-relapser as high-risk), so a specificity-favoring operating point is the defensible choice. *[Suggested cites — confirm fit before using: Vickers & Elkin (2006), "Decision curve analysis: a novel method for evaluating prediction models," Med Decis Making 26(6):565–574 — formalizes choosing the operating point by the FP:FN cost ratio (the direct support for cost-weighting toward specificity); Wynants et al. (2019), "Three myths about risk thresholds for prediction models," BMC Med 17:192 — thresholds should reflect error costs, not be arbitrary; Pepe (2003), The Statistical Evaluation of Medical Tests — foundational sens/spec reference. SUD domain context: Acion et al. (2017), "Use of a machine learning framework to predict substance use disorder treatment success," PLoS ONE 12(4):e0175383. CAVEAT: the SUD ML literature is split — some opioid-relapse/dropout models deliberately prioritize SENSITIVITY (don't-miss-a-relapser); be ready to justify the specificity choice against that.]*

- **Implementing the specificity priority:** AUC tuning is neutral, and the current sens/spec are reported at the **Youden-J cutoff, which weights sensitivity and specificity equally** — so neither step yet encodes this preference. To prioritize specificity, apply a cost function at the operating point: either (a) **re-tune** hyperparameters on a cost-sensitive objective (heavier; the thresholded objective can destabilize selection), or (b) **more cost-effectively, a post-hoc threshold shift** on the same AUC-selected model (cheap, no re-tuning). Prefer (b).

- Word-choice caveat: this is a *prognostic relapse-risk* model, not a *treatment-benefit* model (treatment arm was excluded), so keep causal "benefit of the drug" language out of the final claims.

:::

### Logistic Regression

Standard binary logistic regression models the log-odds of relapse as a linear combination of predictors. Used as the interpretable baseline. Estimated via maximum likelihood; no regularization. Implemented in `tidymodels` using the `glm` engine. Two variants are reported: the full-predictor baseline, and a "no problem variables" model that additionally removes predictors exhibiting complete separation (sparse drug-use indicators whose maximum-likelihood estimate does not exist). Because ordinary logistic regression has no tuning parameters, generalization is estimated by 5-fold cross-validation with `fit_resamples()`, optimizing for — and reporting — cross-validated ROC AUC.

### LASSO and Elastic Net

Penalized logistic regression was fitted with the `glmnet` engine. Two specifications are reported. The elastic-net model tunes both the penalty ($\lambda$) and the mixture (the L1/L2 blend) over a regular 50 × 5 grid via 5-fold cross-validation; the 5-fold cross-validated search selected `mixture = 0.25` (25% L1 / 75% L2 blend — predominantly ridge with a small LASSO component). The pure-LASSO model fixes `mixture = 1` and tunes the penalty alone over 50 log-spaced values. Both optimize ROC AUC, and the final penalty is chosen by the one-standard-error rule (`select_by_one_std_err`), which takes the most heavily penalized model whose CV ROC AUC is within one standard error of the best — favoring sparsity. The L1 penalty is the natural tool for the separation-prone predictors above: it shrinks unstable coefficients toward (elastic net) or exactly to zero (LASSO at the selected penalty).

### K-Nearest Neighbors

K-nearest-neighbors classification was fitted with the `kknn` engine on the standardized predictor matrix (normalization is essential, since KNN uses Euclidean distance). The number of neighbors was tuned over $k = 5, 10, \ldots, 50$ by 5-fold cross-validation optimizing ROC AUC, with the neighbor count chosen by the one-standard-error rule. KNN was run on the full predictor set (including the separation-prone variables): unlike unpenalized logistic regression it has no coefficients to diverge, so it tolerates those predictors directly.

### Planned Tree-Based Models (Future Work)

Classification and regression trees (`rpart`), random forests (`ranger`), and gradient-boosted trees (`xgboost`) are planned extensions. They are not yet implemented and are excluded from the results below, which cover the logistic, penalized-logistic, and KNN models.

## Software

Analyses were conducted in `r stringr::word(R.Version()$version.string, 1, 3)`. Data were preprocessed and visualized with the `tidyverse` (version `r packageVersion("tidyverse")`) [@tidyverse2019] and harmonized clinical-trial data were obtained from `public.ctn0094data` (version `r packageVersion("public.ctn0094data")`). Modeling used the tidymodels ecosystem (version `r packageVersion("tidymodels")`) [@R-tidymodels], including `recipes` for preprocessing, `parsnip` for model specification, `rsample` for resampling, `tune` for hyperparameter search, `yardstick` for metrics, and `workflows` to bind them together. Penalized logistic regression used the `glmnet` engine [@R-glmnet], k-nearest-neighbors used `kknn` [@R-kknn], variable importance used `vip` [@R-vip], and summary tables used `gtsummary` (version `r packageVersion("gtsummary")`) [@R-gtsummary].

# Results

## Participants

```{r participant-counts}
#| include: false
n_by_study <- analysis_base |>
  dplyr::count(project) |>
  dplyr::mutate(project = as.character(project))
get_n <- function(study) n_by_study$n[n_by_study$project == study]
```

After restricting to first randomizations, the analysis sample comprised `r nrow(analysis_base) |> scales::comma()` participants: `r get_n("CTN-0027") |> scales::comma()` from CTN-0027, `r get_n("CTN-0030") |> scales::comma()` from CTN-0030, and `r get_n("CTN-0051") |> scales::comma()` from CTN-0051.

## Participant Characteristics

```{r tbl-demographics}
#| tbl-cap: !expr cap_tbl_demographics

analysis_tibble |>
  select(project, age, is_male, race, is_hispanic,
         education, is_living_stable) |>
  mutate(
    is_male         = factor(is_male,         levels = c(1, 0), labels = c("Yes", "No")),
    is_hispanic     = factor(is_hispanic,     levels = c(1, 0), labels = c("Yes", "No")),
    is_living_stable = factor(is_living_stable, levels = c(1, 0), labels = c("Yes", "No"))
  ) |>
  tbl_summary(
    by = project,
    include = c(age, is_male, race, is_hispanic, education, is_living_stable),
    label = list(
      age              ~ "Age (years)",
      is_male          ~ "Male",
      race             ~ "Race",
      is_hispanic      ~ "Hispanic ethnicity",
      education        ~ "Education",
      is_living_stable ~ "Stable housing"
    ),
    missing = "no"
  ) |>
  modify_header(label = "") |>
  bold_labels()
```

## Drug Use in the Pre-Treatment Window

```{r tbl-drug-prevalence}
#| tbl-cap: !expr cap_tbl_drug

n_analysis <- nrow(analysis_base)

all_drugs_filtered |>
  semi_join(analysis_base, by = "who") |>
  distinct(who, what_grouped) |>
  group_by(what_grouped) |>
  summarize(n_users = n(), .groups = "drop") |>
  mutate(
    pct = round(n_users / n_analysis * 100, 1),
    what_grouped = as.character(what_grouped)
  ) |>
  arrange(desc(n_users)) |>
  kable(
    col.names = c("Substance category", "N users", "% of sample"),
    align = "lrr"
  )
```

## Predictor Distributions

```{r fig-withdrawal-severity}
#| fig-cap: !expr cap_fig_withdrawal
#| fig-width: 6.5
#| fig-height: 4.0

analysis_tibble |>
  filter(!is.na(withdrawal_pre_score)) |>
  mutate(
    severity = factor(
      withdrawal_pre_score,
      levels = 0:3,
      labels = c("None (0)", "Mild (1)", "Moderate (2)", "Severe (3)")
    ),
    project = as.character(project)
  ) |>
  count(project, severity) |>
  group_by(project) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = severity, y = pct, fill = project)) +  # enriched 6/2
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title    = "Pre-induction withdrawal severity by study",
    subtitle = "Percent of each study's participants at each COWS/SOWS severity level",
    x = "Pre-induction withdrawal severity",
    y = "Percentage of participants (%)",
    fill = "Study"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

## Feature Matrix

```{r feature-dimensions}
n_participants <- nrow(analysis_tibble)
n_features     <- ncol(analysis_tibble) - 2L   # subtract who and project
```

The analysis tibble contains **`r n_participants |> scales::comma()` participants** and **`r n_features` engineered feature columns** (excluding participant ID and study assignment), spanning substance-use history (per-drug days, binary use, longest streak), polydrug days, drug-use breadth, demographics, addiction severity (ASI), nicotine dependence (Fagerström), pain, psychiatric comorbidity, quality of life, housing stability, IV-use risk behavior, withdrawal severity, prescribed-medication adherence, study site, and cross-domain interactions. This matrix is shipped as the `analysis_tibble` dataset in the `ctn0094.oliven` package.

## Logistic Regression

```{r tbl-lr-or}
#| tbl-cap: "Logistic regression odds ratios (p < 0.05, sorted by effect size). Wald 95% confidence intervals."

lr_or_table |>
  kable(
    digits    = 3,
    col.names = c("Predictor", "Odds Ratio", "95% CI Low", "95% CI High", "p-value")
  )
```

```{r tbl-lr-metrics}
#| tbl-cap: "Logistic regression model performance. Training and test sets from a stratified 75/25 split."

bind_cols(
  lr_metrics      |> rename(Train = .estimate),
  lr_test_metrics |> select(Test = .estimate)
) |>
  mutate(.metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) |>
  kable(
    digits    = 3,
    col.names = c("Metric", "Train", "Test")
  )
```

### Exclusion of Structurally Sparse Predictors

```{r wopv-ci-check}
#| include: false
wopv_ci_check <- lr_fit_wopv |>
  tidy(exponentiate = TRUE) |>
  mutate(
    conf.low    = exp(log(estimate) - 1.96 * std.error),
    conf.high   = exp(log(estimate) + 1.96 * std.error),
    log10_width = log10(conf.high / conf.low)
  ) |>
  filter(!is.na(estimate), term != "(Intercept)")

wopv_top3_ci <- wopv_ci_check |> slice_max(log10_width, n = 3)
```

A second logistic regression was fitted after excluding predictors identified as problematic on substantive or numerical grounds. Five predictors were excluded for leakage: `took_own_study_drug`, `took_other_study_drug`, `methadone_binary`, `buprenorphine_binary`, and `suboxone_binary` encode treatment arm assignment directly or by proxy, introducing post-randomization information into the pre-treatment predictor set.

The remaining exclusions addressed complete separation. Inspection of the baseline coefficient table revealed predictors with Wald confidence intervals spanning 25–90 orders of magnitude — the diagnostic signature of a non-existent maximum likelihood estimate. When a predictor nearly perfectly partitions the binary outcome, the log-likelihood has no finite maximum and the iteratively reweighted least squares optimizer terminates at an arbitrarily large value; the resulting point estimate is numerically meaningless regardless of its apparent plausibility. All such predictors were sparse drug-use indicators (analgesic, antiemetic, and muscle relaxant use and their binary counterparts) with very few observed events in the 28-day pre-treatment window.

Exclusion proceeded over three iterative rounds. In each round, all predictors with confidence intervals spanning more than approximately 10 orders of magnitude were removed, the model was refitted, and the coefficient table was re-examined. Predictors appearing in subsequent rounds arose exclusively from correlated pairs (Pearson *r* > 0.90 at the `step_corr` threshold): removing the first member of a pair permitted the second to pass the pairwise correlation filter, where it exhibited the same separation pattern. This sequential structure confirms that the final stable configuration reflects genuine structural sparsity rather than an incidental local solution — every excluded variable was either directly sparse or a downstream consequence of direct sparsity.

Following three rounds of exclusion totaling 15 predictors (beyond `treatment_arm`, which is excluded from both models), all `r nrow(wopv_ci_check)` estimable predictors had well-bounded Wald confidence intervals. The three widest spanned `r round(wopv_top3_ci$log10_width[1], 1)`, `r round(wopv_top3_ci$log10_width[2], 1)`, and `r round(wopv_top3_ci$log10_width[3], 1)` orders of magnitude (`r wopv_top3_ci$term[1]`, `r wopv_top3_ci$term[2]`, and `r wopv_top3_ci$term[3]`, respectively); all remaining intervals were narrower. Seven additional terms had missing coefficient estimates due to exact multicollinearity in the encoded design matrix (rank deficiency); these reflect structural properties of the one-hot and polynomial encodings and are not separation.

```{r tbl-lr-wopv-metrics}
#| tbl-cap: "Logistic regression (glm, no regularization) performance after exclusion of structurally sparse predictors (SSPs). Training and test sets from a stratified 75/25 split (seed 12345). No cross-validation — the model has no tuning parameters; 5-fold CV performance is shown separately in the all-methods comparison table."

bind_cols(
  lr_metrics_wopv      |> rename(Train = .estimate),
  lr_test_metrics_wopv |> select(Test = .estimate)
) |>
  mutate(.metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) |>
  kable(
    digits    = 3,
    col.names = c("Metric", "Train", "Test")
  )
```

## Model Performance Across Methods

```{r tbl-all-metrics}
#| tbl-cap: "ROC AUC for all fitted models across three evaluation contexts: 5-fold cross-validation on the training set (used for hyperparameter selection; logistic CV uses fit_resamples for performance estimation), full in-sample training set AUC, and held-out 25% test set AUC. Care-seeking adults with opioid use disorder in CTN-0027/0030/0051 (2006-2016), N = 2,492."

# Helper to extract ROC AUC from a .metric/.estimate tibble
get_auc <- function(tbl) round(tbl$.estimate[tbl$.metric == "roc_auc"], 3)

# Read from named objects in logistic_enet.R. # added 6/2
enet_cv_auc     <- enet_cv_metrics$mean[enet_cv_metrics$.metric == "roc_auc"]
lasso_cv_auc    <- lasso_cv_metrics$mean[lasso_cv_metrics$.metric == "roc_auc"]
enet_train_auc  <- enet_train_metrics$.estimate[enet_train_metrics$.metric == "roc_auc"]
lasso_train_auc <- lasso_train_metrics$.estimate[lasso_train_metrics$.metric == "roc_auc"]
enet_test_auc   <- enet_test_metrics$.estimate[enet_test_metrics$.metric == "roc_auc"]
lasso_test_auc  <- lasso_test_metrics$.estimate[lasso_test_metrics$.metric == "roc_auc"]
# Read from named objects in cart.R.
cart_cv_auc    <- cart_cv_metrics$mean[cart_cv_metrics$.metric == "roc_auc"]
cart_train_auc <- cart_train_metrics$.estimate[cart_train_metrics$.metric == "roc_auc"]
cart_test_auc  <- cart_test_metrics$.estimate[cart_test_metrics$.metric == "roc_auc"]
lr_cv_auc <- lr_cv_test_metrics$.estimate[lr_cv_test_metrics$.metric == "roc_auc"]

# Full training AUC (in-sample, on all training data after final fit)
lr_train_auc      <- get_auc(lr_metrics)
lr_wopv_train_auc <- get_auc(lr_metrics_wopv)

# Test AUC
lr_test_auc      <- get_auc(lr_test_metrics)
lr_wopv_test_auc <- get_auc(lr_test_metrics_wopv)
knn_test_auc     <- knn_test_metrics$.estimate[knn_test_metrics$.metric == "roc_auc"]
knn_wopv_test_auc <- knn_wopv_test_metrics$.estimate[knn_wopv_test_metrics$.metric == "roc_auc"]

tibble::tribble(
  ~Method,                         ~CV,                        ~Train,           ~Test,
  "Logistic (all predictors)",     NA_real_,                   lr_train_auc,      lr_test_auc,
  "Logistic (no problem vars)",    NA_real_,                   lr_wopv_train_auc, lr_wopv_test_auc,
  "Logistic (5-fold CV)",          round(lr_cv_auc, 3),        NA_real_,          NA_real_,
  "Elastic net (mixture = 0.25)",  round(enet_cv_auc, 3),      round(enet_train_auc, 3),  enet_test_auc,
  "LASSO",                         round(lasso_cv_auc, 3),     round(lasso_train_auc, 3), lasso_test_auc,
  "KNN",                           round(knn_cv_auc, 3),       round(knn_train_auc, 3),   knn_test_auc,
  "KNN (no problem vars)",         round(knn_wopv_cv_auc, 3),  round(knn_wopv_train_auc, 3), knn_wopv_test_auc,
  "CART",                          round(cart_cv_auc, 3),      round(cart_train_auc, 3),     cart_test_auc
) |>
  kable(
    digits    = 3,
    col.names = c("Method", "CV AUC", "Full Train AUC", "Test AUC"),
    align     = "lrrr",
    na        = "—"
  )
```

The full-predictor and no-problem-variable logistic models perform near-identically on test ROC AUC (`r lr_test_auc` vs `r lr_wopv_test_auc`), so excluding the separation-prone predictors costs essentially no discrimination while removing numerically meaningless coefficients. We therefore consider the no-problem-variable logistic model the most trustworthy linear specification for handling separation. KNN and LASSO, by contrast, are run on the full predictor set: their built-in mechanisms (distance-based prediction and L1 penalization, respectively) absorb the sparse, separating variables without the coefficient blow-up that destabilizes unpenalized logistic regression.

## Variable Importance by Engine

```{r fig-vip-glm}
#| fig-cap: !expr cap_vip_glm
#| fig-width: 6.5
#| fig-height: 4.5
vip::vip(lr_wopv_vip_obj, num_features = 15, geom = "col") +
  ggplot2::labs(title = NULL, x = "Importance (|z-statistic|)", y = NULL) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r fig-vip-glmnet}
#| fig-cap: !expr cap_vip_glmnet
#| fig-width: 6.5
#| fig-height: 4.5
ggplot2::ggplot(lasso_coef,
  ggplot2::aes(x = importance, y = reorder(term, importance))) +
  ggplot2::geom_col(fill = "grey40") +
  ggplot2::labs(x = "Importance (|coefficient| at selected lambda)", y = NULL) +
  ggplot2::theme_minimal(base_size = 11)
```

```{r fig-vip-knn}
#| fig-cap: !expr cap_vip_knn
#| fig-width: 6.5
#| fig-height: 4.5
vip::vip(knn_vip_obj, num_features = 15, geom = "col") +
  ggplot2::labs(x = "Importance (mean ROC AUC decrease)", y = NULL) +
  ggplot2::theme_minimal(base_size = 11)
```

# Discussion

The feature matrix described here represents the pre-treatment behavioral and clinical profile for each randomized participant across the three CTN studies. The full feature set spans substance use, demographics, psychiatric comorbidity, pain, nicotine dependence, social risk behavior, IV drug use, withdrawal severity, medication adherence, and study site. These features will be used to train and evaluate predictive models of relapse, with the aim of informing treatment matching for OUD.

The preprocessing recipe handles two distinct classes of missingness separately: (1) structural MNAR missingness, where entire studies did not collect certain variables (e.g., substance use disorder diagnosis flags, homeless status), addressed by creating an explicit is-missing indicator column via `step_indicate_na()` before imputing, so the model can learn from the missingness pattern itself; and (2) random MCAR missingness, affecting a small number of participants across variables such as pain score and withdrawal severity, addressed by median or mode imputation with no indicator. Additional preprocessing steps include `step_ordinalscore()` for ordered factors with defensible equal spacing, polynomial contrast encoding for cigarettes-per-day, and `step_corr()` to remove predictors with pairwise correlation above 0.90, which eliminates redundant pairs such as the two withdrawal severity columns derived from the same single observation.

The baseline logistic regression achieved a ROC AUC of `r round(lr_metrics$.estimate[lr_metrics$.metric == "roc_auc"], 3)` on training and `r round(lr_test_metrics$.estimate[lr_test_metrics$.metric == "roc_auc"], 3)` on test — a mild to moderate drop consistent with overfitting expected from fitting an unregularized model to 100+ predictors, but still reflecting meaningful discriminative ability above chance. The large estimated associations for study membership (CTN-0030: OR ≈ 42, CTN-0027: OR ≈ 4 in the cleaned model) likely reflect between-study heterogeneity in outcome ascertainment or follow-up design rather than pre-treatment clinical effects, consistent with the challenge of pooling trials with distinct protocols.

## Strengths and Weaknesses by Method

**Logistic regression (glm).** *Strengths:* fully interpretable (odds ratios with confidence intervals), no tuning required, and provides an honest probabilistic baseline. *Weaknesses:* with 100+ predictors and several near-separating sparse indicators, the unpenalized fit is numerically unstable (coefficients and confidence intervals diverge), and it overfits, as seen in the train-to-test ROC AUC drop. Removing the problematic predictors restores stable estimates at essentially no cost to discrimination.

**LASSO / elastic net (glmnet).** *Strengths:* built-in regularization tames the separation-prone predictors and performs automatic feature selection, yielding a sparse, deployable model from the full predictor set; the one-standard-error rule guards against overfitting. *Weaknesses:* penalized coefficients are biased toward zero (shrinkage trades a little unbiasedness for stability), the selected sparse set can be unstable across resampling folds, and standardized penalized coefficients are less directly interpretable as odds ratios than unpenalized estimates.

**K-nearest neighbors (kknn).** *Strengths:* non-parametric and able to capture local, non-linear structure with a single tuning parameter; tolerates the separating predictors without coefficient blow-up. *Weaknesses:* no intrinsic variable importance (we resort to computationally expensive permutation importance), sensitive to the curse of dimensionality with 100+ standardized features, exposes no interpretable model form, and prediction cost grows linearly with the training set.

Across methods, test ROC AUC is similar, indicating that the linear signal is modest and not dramatically improved by KNN's non-parametric flexibility; model choice here is therefore driven as much by interpretability and stability as by discrimination.

## LASSO Sparsity and Predictive Performance

A notable feature of the pure LASSO model is that it retains meaningful predictive performance on a very sparse variable set. Of the approximately 107 candidate predictors entering the penalized fit, only `r nrow(lasso_coef)` survive with non-zero coefficients at the selected penalty (chosen by the 1-SE rule). The surviving predictors are:

```{r lasso-surviving-vars}
#| echo: false
lasso_coef |>
  dplyr::arrange(dplyr::desc(importance)) |>
  dplyr::pull(term) |>
  paste(collapse = ", ") |>
  cat()
```

Despite discarding `r 107 - nrow(lasso_coef)` predictors entirely, the LASSO test ROC AUC (`r round(lasso_test_auc, 3)`) is within `r round(abs(lasso_test_auc - lr_wopv_test_auc), 3)` AUC points of the unpenalized no-problem-variables logistic regression — a negligible practical difference. This is consistent with the predictive signal being diffuse and redundant across many correlated predictors: LASSO picks one representative from each correlated cluster and zeros the rest, so the discarded variables are not unimportant in isolation — they are redundant *given* the survivors.

The sparse surviving set should therefore be interpreted cautiously on two grounds. First, LASSO variable selection is unstable under correlated predictors: a different random seed or fold assignment would likely yield a substantially different sparse set with essentially the same AUC, because the selection is navigating a flat ridge in the likelihood surface rather than identifying a unique minimal predictor set. Second, the surviving variables may not correspond to the clinically most meaningful predictors — LASSO selects whichever variable from a correlated cluster has marginally better alignment with the outcome in this specific training sample, which can make the selected set feel arbitrary. The near-identical AUC between sparse and full models is more a symptom of weak, diffuse signal and high predictor redundancy than evidence that these particular variables are the true drivers of relapse.

```{r fig-vip-lasso-discussion}
#| fig-cap: !expr cap_vip_glmnet
#| fig-width: 6.5
#| fig-height: 4.5
ggplot2::ggplot(lasso_coef,
  ggplot2::aes(x = importance, y = reorder(term, importance))) +
  ggplot2::geom_col(fill = "grey40") +
  ggplot2::labs(
    title    = "LASSO surviving predictors at the selected penalty",
    subtitle = paste0(nrow(lasso_coef), " non-zero coefficients of ~107 total"),
    x        = "Importance (|coefficient| at selected lambda)",
    y        = NULL
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

## Study Membership as a Predictor

Study membership (`project_X30` and `project_X27` in the one-hot encoded feature matrix — with `project_X51` as the reference level) appears among the stronger predictors in the logistic regression (OR ≈ 42 for CTN-0030, OR ≈ 4 for CTN-0027 relative to CTN-0051). This warrants explanation.

Study membership is **not leakage**: it is known at enrollment, before any treatment is assigned or administered, and it does not encode which treatment a participant received within a study (that is `treatment_arm`, which was explicitly excluded from all models). It is legitimate predictor information in the same way that enrolling site or referral source would be.

*Interpretation (speculative):* The large study effects likely reflect at least three overlapping mechanisms. First, and probably dominant, each trial enrolled a distinct **population**: eligibility criteria, geographic sites, time periods, and demographic compositions differed across CTN-0027, -0030, and -0051, and those population-level differences translate into genuine between-study differences in baseline relapse risk. Second, each study carried a distinct **protocol context** — infrastructure, follow-up schedule, and clinical staffing — that is non-leaky and may itself influence outcome. Note that this is distinct from the specific treatment a participant received within the study: participants were randomized to different arms within each trial, so `project` does not encode which drug or regimen a given participant got (that is `treatment_arm`, excluded). Third, the harmonized outcome variable `lee2018_rel_event` was constructed from studies using different assessment instruments (COWS vs. SOWS), follow-up windows, and ascertainment methods; some of the project effect may reflect **outcome measurement heterogeneity** across protocols rather than true differences in relapse. This last point is a limitation of pooled-trial analysis generally.

*Generalizability caveat (speculative):* Because `project` is a coarse proxy for all of the above, the model as trained may not generalize well to new studies with different populations or protocols. A more generalizable version would encode features of the population and treatment setting directly (e.g., eligibility criteria, medication type, treatment duration, site support resources) rather than relying on study-membership dummies — replacing the proxy with its constituents.

# References {.unnumbered}

```{r make-references}
#| include: false

knitr::write_bib(
  c(.packages(), "rUM", "glmnet", "kknn", "vip", "tidymodels"),
  "packages.bib"
)
```
