---
title: "Predicting treatment response for people undergoing care for OUD"
author: "Nat Oliven"
date: last-modified
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predicting treatment response for people undergoing care for OUD}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    echo: false
    message: false
    warning: false
    error: false
    comment: ""
    dpi: 100
    fig-width: 6.5
    fig-height: 4.0
    R.options:
      digits: 3
editor: source
bibliography: [references.bib, packages.bib]
csl: apa.csl
---

```{r load-packages}
library(conflicted)
suppressPackageStartupMessages(library(tidymodels))
tidymodels_prefer()
suppressPackageStartupMessages(library(tidyverse))
library(public.ctn0094data)
library(gtsummary)
library(knitr)
conflicts_prefer(dplyr::filter)
options(dplyr.summarise.inform = FALSE)
```

```{r build-analysis}
#| cache: true
#| include: false
source("building_analysis.R")
```

# Abstract

This study uses data from three NIDA Clinical Trials Network studies (CTN-0028, CTN-0030, CTN-0051) to build a predictive model of treatment response in people with opioid use disorder (OUD). We construct a pre-treatment feature matrix from behavioral, substance use, demographic, and clinical data in the 28 days before randomization, and evaluate whether these features predict relapse during the treatment period.

# Introduction

Opioid use disorder is associated with substantial mortality, morbidity, and economic burden in the United States. Medications for OUD (MOUD) — including methadone, buprenorphine-naloxone, and extended-release naltrexone — are effective treatments, yet individual response varies considerably. Matching patients to the treatment most likely to benefit them remains a central challenge in addiction medicine.

The CTN-0094 harmonized dataset combines three randomized clinical trials comparing MOUD regimens in opioid-dependent adults. By pooling data across studies, we have sufficient sample size to examine whether pre-treatment characteristics predict treatment response. The goal of this analysis is to build and evaluate a feature matrix capturing the 28 days prior to randomization, which will serve as input to predictive models of relapse outcomes.

# Methods

## Data

Data come from three NIDA Clinical Trials Network studies harmonized in the `public.ctn0094data` R package: CTN-0028 (methadone vs. buprenorphine-naloxone; n = 1,920), CTN-0030 (buprenorphine-naloxone only; n = 868), and CTN-0051 (extended-release naltrexone vs. buprenorphine-naloxone; n = 772). Only participants' first randomization is included; the CTN-0030 re-randomization step is excluded via an inner join on the earliest randomization date per participant.

## Feature Engineering

All longitudinal features are derived from a 28-day pre-randomization window (days −28 inclusive to day 0 exclusive), providing exactly four of each weekday per participant. Substance use features are computed from the `all_drugs` dataset, which is more granular than the standalone `tlfb` table and covers multiple detection sources (TLFB self-report, urine drug screen, and urine drug screen with alcohol biomarker). Rare drug categories (<10 events in the window) and illicit categories used by fewer than 1% of participants are excluded before feature computation. Per-drug features include days of use, binary use indicator, and longest consecutive streak where clinically meaningful. Additional feature domains include demographics, self-reported behavior risk, addiction severity, psychiatric comorbidity, quality of life, pain, nicotine dependence, and withdrawal severity. Feature engineering is implemented in `building_analysis.R`; sections requiring dataset column name resolution are marked with TODO and return placeholder columns until resolved.

Analyses were conducted with `r stringr::word(R.Version()$version.string, 1, 3)` with the `tidyverse` (`r packageVersion("tidyverse")`), `tidymodels` (`r packageVersion("tidymodels")`), and `gtsummary` (`r packageVersion("gtsummary")`) packages. [@R-base; @R-tidyverse; @tidyverse2019; @R-gtsummary]

# Results

## Participants

```{r tbl-participants}
#| tbl-cap: "Randomized participants by CTN study."

analysis_base |>
  count(project) |>
  mutate(project = as.character(project)) |>
  kable(col.names = c("Study", "N"), align = "lr")
```

A total of `r nrow(analysis_base)` participants across `r n_distinct(analysis_base$project)` studies were included after restricting to first randomizations (@tbl-participants).

## Participant Characteristics

```{r tbl-demographics}
#| tbl-cap: "Baseline participant characteristics by study."

analysis_tibble |>
  select(project, age, is_male, race, is_hispanic,
         education, is_living_stable) |>
  mutate(
    is_male         = factor(is_male,         levels = c(1, 0), labels = c("Yes", "No")),
    is_hispanic     = factor(is_hispanic,     levels = c(1, 0), labels = c("Yes", "No")),
    is_living_stable = factor(is_living_stable, levels = c(1, 0), labels = c("Yes", "No"))
  ) |>
  tbl_summary(
    by = project,
    include = c(age, is_male, race, is_hispanic, education, is_living_stable),
    label = list(
      age              ~ "Age (years)",
      is_male          ~ "Male",
      race             ~ "Race",
      is_hispanic      ~ "Hispanic ethnicity",
      education        ~ "Education",
      is_living_stable ~ "Stable housing"
    ),
    missing = "no"
  ) |>
  modify_header(label = "") |>
  bold_labels()
```

## Drug Use in the Pre-Treatment Window

```{r tbl-drug-prevalence}
#| tbl-cap: "Substance use categories observed in the 28 days before randomization (days −28 to −1). N users and percentage are relative to the analysis sample."

n_analysis <- nrow(analysis_base)

all_drugs_filtered |>
  semi_join(analysis_base, by = "who") |>
  distinct(who, what_grouped) |>
  group_by(what_grouped) |>
  summarize(n_users = n(), .groups = "drop") |>
  mutate(
    pct = round(n_users / n_analysis * 100, 1),
    what_grouped = as.character(what_grouped)
  ) |>
  arrange(desc(n_users)) |>
  kable(
    col.names = c("Substance category", "N users", "% of sample"),
    align = "lrr"
  )
```

## Feature Matrix

```{r feature-dimensions}
n_participants <- nrow(analysis_tibble)
n_features     <- ncol(analysis_tibble) - 2L   # subtract who and project
```

The current feature matrix contains **`r n_participants` participants** and **`r n_features` feature columns** (excluding participant ID and study). Feature domains that depend on unresolved dataset column names (ASI composites, Fagerstrom score, pain scores, psychiatric diagnoses, quality of life, treatment arm, risk behavior, and withdrawal trajectory) contribute placeholder columns and will be expanded once column names are confirmed via the `glimpse()` calls in `building_analysis.R`.

# Discussion

The feature matrix described here represents the pre-treatment behavioral and clinical profile for each randomized participant across the three CTN studies. Once the remaining TODO sections are resolved, the full feature set will include substance use history, demographic characteristics, psychiatric comorbidity, pain, nicotine dependence, social risk behavior, IV drug use, withdrawal severity, medication adherence, and study site. These features will be used to train and evaluate predictive models of relapse, with the aim of informing treatment matching for OUD.

# References {.unnumbered}

```{r make-references}
#| include: false

knitr::write_bib(
  c(.packages(), "rUM"),
  "packages.bib"
)
```
