---
title: "Features to Include — Analysis-Ordered"
author: "Nat Oliven"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

<!--
HEADING CONVENTIONS (from Features_To_Include_Accepted_Suggestions.Rmd — retained for reference, no longer binding in this document)

# Formatting Conventions For This Document

- **Header 1's** are datasets, and "Database Notes" about the data package as a whole
- **Header 2's** are variables within the datasets, and "Dataset Notes" about that dataset as a whole (can include photos of a table, descriptive graphics, anything that gives us a sense of what we're looking at)
  - "Who" is a redundant column that we don't need to investigate every time, include variables of note
  - Ideally we'd have all 46 variables, but maybe the less interesting ones don't get explored.
- **Header 3's** are always written "Feature proposal", with text underneath explaining what the feature is and why it should be included.
  - When a "Feature proposal" pertains to just one variable in the dataset, put it beneath that variable's "Header 2"
    - If the raw column itself is a feature proposal, that's ok! Just write this explicitly and explain why it requires no further filtering
  - When a "Feature proposal" pertains to multiple variables in the same dataset, put the feature proposal in "Dataset Notes" for that dataset ("Dataset Notes" will be a Header 2)
    - When the "Feature proposal" pertains to multiple variables in different datasets, put the feature proposal in the "Database Notes"
- To make headers, highlight your text, and then go to Format > Paragraph Styles and select the appropriate header.
-->

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(conflicted)
suppressPackageStartupMessages(library(tidyverse))
library(public.ctn0094data)
conflicts_prefer(dplyr::filter)
```

## Data Set Description

Data sets in package `public.ctn0094data`:

| Dataset                     | Description                                            |
|-----------------------------|--------------------------------------------------------|
| `all_drugs`                 | All drugs taken                                        |
| `asi`                       | Did patient use intravenous drugs                      |
| `demographics`              | Patient demographics                                   |
| `detox`                     | Start and Stop of Detox                                |
| `everybody`                 | Everybody with any data                                |
| `fagerstrom`                | Fagerstrom Test for Nicotine Dependence                |
| `first_survey`              | First Survey Date                                      |
| `meta_study_length`         | Metadata About Study Length                            |
| `meta_substance_groups_uds` | Metadata About UDS Groupings                           |
| `pain`                      | Self-Reported Pain                                     |
| `psychiatric`               | Psychiatric History                                    |
| `qol`                       | Quality of Life                                        |
| `randomization`             | Randomization Data                                     |
| `rbs`                       | Risk Behavior Survey                                   |
| `rbs_iv`                    | Risk Behavior Survey IV drug use information           |
| `screening_date`            | Screening Date Information                             |
| `sex`                       | Sexual Activity in Risk Behavior Survey                |
| `site_masked`               | Site Regrouped                                         |
| `tlfb`                      | Timeline Followback (TLFB) Drug Use Information        |
| `treatment`                 | Amount of Study Drug Per Day                           |
| `uds`                       | Urine Drug Screening (UDS) Results                     |
| `uds_temp`                  | Urine Drug Screening (UDS) Temperature                 |
| `visit`                     | Patient Visit Data                                     |
| `withdrawal`                | Patient Withdrawal Symptoms Per Day                    |
| `withdrawal_pre_post`       | Patient Withdrawal Symptoms Pre and Post Induction     |

---

# ── 1. Base tibble: first randomization only ──────────────────────

## randomization

Note: make sure to include first randomization only.

## everybody

### who

Description: Ascending ID's from 1 to 3560 for the 3560 participants. They're all unique, nothing messy here.

Notes: We will commonly join on `who` to compare info across datasets

### project

Description: Three possible values: 27, 30, 51, corresponding to the

```{r project-levels}
everybody |> distinct(project)
```

Notes: We can get a lot of exploratory information by separating out which study the participants were on. For example, study 27 has very sparse demographic information in demographics, across lots of the variables.

### Dataset Notes

Number of people on each study

```{r project-counts}
everybody |>
  group_by(project) |>
  summarize(count = n())
```

---

# ── 2. all_drugs setup ────────────────────────────────────────────

## all_drugs

Rather than using tlfb, I cleaned up all_drugs to give me more granular control over the rule. I axed any meds with less than 10 uses. I grouped together all opioids. I left out certain illicit drugs with significance in the literature related to opioid relapse (fent, methamphetamines, cocaine, etc.) separate. I kept hard and light drinking separate.

```{r all-drugs-map}
drug_map <- c(
  # opioid group, but keep heroin/fentanyl/buprenorphine/suboxone separate
  "Acetaminophen"          = "Analgesic",
  "Codeine"                = "Opioid",
  "Hydrocodone"            = "Opioid",
  "Hydromorphone"          = "Opioid",
  "Merperidine"            = "Opioid",
  "Morphine"               = "Opioid",
  "Nalbuphine"             = "Opioid",
  "Opium"                  = "Opioid",
  "Oxycodone"              = "Opioid",
  "Oxymorphone"            = "Opioid",
  "Propoxyphene"           = "Opioid",
  "Tramadol"               = "Opioid",

  # sedatives, but keep benzodiazepines separate
  "Sedative-Hypnotic"      = "Sedatives",
  "Barbiturate"            = "Sedatives",

  # antidepressants
  "Trazodone"              = "Antidepressant",
  "Tryclic-Antidepressant" = "Antidepressant",

  # analgesics
  "Gabapentin"             = "Analgesic",

  # naming / grouping cleanup
  "Thc"                    = "Cannabinoids",
  "K2"                     = "Cannabinoids",
  "Musclerelax"            = "Muscle Relaxant",
  "Mdma"                   = "MDMA/Hallucinogen",
  "Hallucinogen"           = "MDMA/Hallucinogen",
  # Not aggregating — renaming only so all alcohol categories sort together alphabetically
  "Heavy Drinking"         = "Alcohol Heavy Amnt",
  "Light Drinking"         = "Alcohol Light Amnt",
  "Alcohol"                = "Alcohol Missing Amnt"
)

all_drugs_grouped <- all_drugs |>
  mutate(
    what_grouped = ifelse(
      what %in% names(drug_map),
      drug_map[as.character(what)],
      as.character(what)
    ),
    what_grouped = factor(what_grouped)
  )
```

```{r all-drugs-filtered}
# Primary filter: keep only categories with >= 10 use events in window
drugs_to_keep <- all_drugs_grouped |>
  filter(when < 0, when >= -28) |>
  count(what_grouped) |>
  filter(n >= 10) |>
  pull(what_grouped)

all_drugs_filtered <- all_drugs_grouped |>
  filter(when < 0, when >= -28) |>
  filter(what_grouped %in% drugs_to_keep)

# Secondary filter: drop illicit categories used by < 1% of
# participants. Prescribed / legal categories are exempt because
# even rare use carries signal (medication adherence, comorbidity).
prescribed_or_legal <- c(
  "Buprenorphine", "Suboxone", "Methadone",
  "Benzodiazepine", "Sedatives", "Opioid",
  "Muscle Relaxant", "Analgesic", "Antidepressant", "Antiemetic",
  "Alcohol Heavy Amnt", "Alcohol Light Amnt", "Alcohol Missing Amnt",
  "Cannabinoids", "Nicotine", "Caffeine"
)

total_persons_all <- n_distinct(all_drugs_filtered$who)

illicit_low_prev <- all_drugs_filtered |>
  distinct(who, what_grouped) |>
  group_by(what_grouped) |>
  summarize(n_users = n(), .groups = "drop") |>
  mutate(pct_users = n_users / total_persons_all * 100) |>
  filter(pct_users < 1,
         !as.character(what_grouped) %in% prescribed_or_legal) |>
  pull(what_grouped)

all_drugs_filtered <- all_drugs_filtered |>
  filter(!what_grouped %in% illicit_low_prev)

all_drugs_filtered |>
  group_by(what_grouped) |>
  count() |>
  arrange(desc(n)) |>
  print(n = 30)
```

#### Feature proposal

Drug grouping validation notes:

1. ~~**⚠️ WARNING — Acetaminophen in the "Opioid" group**~~ **✅ RESOLVED**: Acetaminophen was moved from the "Opioid" group to the "Analgesic" group in `building_analysis.R` (`drug_map`). It is now grouped with Gabapentin as `Analgesic`, preventing double-counting against `opioid_days`. See `drug_specific_features.md` → Analgesic section for the rationale.

2. ~~**CONSIDER — Gabapentin standalone instead of "Analgesic"**~~ **✅ DECIDED — keep as "Analgesic"**: Gabapentin remains grouped with Acetaminophen under the `Analgesic` category in `building_analysis.R`. Rationale: Acetaminophen alone has only 5 events in the window and fails the ≥10 primary filter; grouping with Gabapentin (23 events) gives a combined count of 28 and allows both to survive. The `analgesic` features are therefore best read as "any prescribed non-opioid analgesic," not as a pure gabapentin misuse signal. The standalone-gabapentin option remains available if the composite signal proves uninformative at modelling time.

### Dataset Notes

---

# ── 3. (Reserved) ────────────────────────────────────────────────

<!-- No source-document content maps here. Nicotine dependence is captured via the Fagerstrom score only (see §11). -->

---

# ── 4. Helper functions ───────────────────────────────────────────

<!-- No source-document content maps here. Technical R implementation only. -->

---

# ── 5. Per-drug features ──────────────────────────────────────────

## all_drugs

### Feature proposals: Days per person

For each drug category below, compute total days in the window −28 to −1 where **any source** (TFB, UDS, or UDSAB) detected the substance. Count each day at most once per drug per person (no double-counting across sources).

Include only drug categories that pass the `>=10 events` filter. Verify the final list by running `all_drugs_filtered |> count(what_grouped)` before building the analysis tibble.

| Feature name | Drug category | Rationale for inclusion |
|---|---|---|
| `heroin_days` | Heroin | Primary opioid of abuse; strongest relapse predictor |
| `fentanyl_days` | Fentanyl | Increasingly prevalent; distinct overdose pharmacology |
| `opioid_days` | Opioid (prescription group) | Pre-study prescription opioid misuse as a composite |
| `cocaine_days` | Cocaine | Stimulant co-use is associated with OUD relapse |
| `crack_days` | Crack cocaine | Distinct use pattern from powder cocaine; include if ≥10 events |
| `methamphetamine_days` | Methamphetamine | Stimulant; increasingly common in OUD populations |
| `benzodiazepine_days` | Benzodiazepine | Cross-addiction; high overdose risk; strong predictor |
| `cannabinoids_days` | Cannabinoids (THC + K2) | Complex relationship with opioid craving; include |
| `alcohol_heavy_amnt_days` | Alcohol Heavy Amnt | Heavy drinking directly predicts opioid relapse |
| `alcohol_light_amnt_days` | Alcohol Light Amnt | Lower risk but captures broader drinking behavior |
| `buprenorphine_days` | Buprenorphine | Pre-study MOUD exposure; see "Took THEIR study drug" feature |
| `suboxone_days` | Suboxone | Treated as buprenorphine-equivalent per Nat's note |
| `methadone_days` | Methadone | Pre-study MOUD exposure |
| `sedatives_days` *(conditional)* | Sedatives | Include if ≥10 events; non-benzo CNS depressants |
| `analgesic_days` *(conditional)* | Analgesic (Gabapentin + Acetaminophen) | Include if ≥10 events; growing misuse concern in OUD |
| `antidepressant_days` *(conditional)* | Antidepressant | Include if ≥10 events; functions as prescription adherence marker |
| ~~`caffeine_days`~~ | Caffeine | **Omit** — no established link to OUD relapse outcomes |
| `nicotine_days` *(conditional)* | Nicotine | Include if ≥10 events; comorbidity (better measured by Fagerstrom, but days count adds information) |

### Feature proposals: Binary (any use)

For each drug in the days table above, also compute a binary 0/1 flag: did the person use this substance **at all** in the −28 to −1 window (any source, any day)?

Binary features are especially useful for models with linearity assumptions — the difference between 0 and 1 days should often count more than the difference between, say, 12 and 13 days.

Also include `nicotine_binary` even if `nicotine_days` is omitted (ever-smoker status adds independent predictive value over the continuous Fagerstrom score).

### Feature proposals: Longest streak

Compute the longest consecutive-day streak for **addictive substances only**. Streak length captures compulsive, dependent use patterns beyond what a total-days count reflects.

**Include streaks for:**

| Feature name | Drug category | Justification |
|---|---|---|
| `heroin_streak` | Heroin | Highly addictive; streak reflects dependence severity |
| `fentanyl_streak` | Fentanyl | Same rationale as heroin |
| `opioid_streak` | Opioid (prescription group) | Compulsive prescription opioid use pattern |
| `cocaine_streak` | Cocaine | Stimulant use disorder; binge pattern captured by streak |
| `crack_streak` | Crack cocaine | Same rationale as cocaine |
| `methamphetamine_streak` | Methamphetamine | High addiction potential; streak is meaningful |
| `benzodiazepine_streak` | Benzodiazepine | Physical dependence develops; streak indicates severity |
| `alcohol_heavy_amnt_streak` | Alcohol Heavy Amnt | Heavy drinking streaks predict relapse more than isolated events |
| `cannabinoids_streak` | Cannabinoids (THC + K2) | Cannabis use disorder is real; streak captures it |
| `sedatives_streak` *(conditional)* | Sedatives | Include if ≥10 events; same logic as benzos |

**Omit streaks for:**

- `buprenorphine`, `suboxone`, `methadone` — these are medications; a long streak is desirable adherence, not compulsive use
- `antidepressant`, `analgesic` — prescription medications; streak ≠ addictive behavior
- `alcohol_light_amnt` — light drinking streaks are less clinically meaningful
- `caffeine` — omitted entirely

---

## Drug-specific feature decisions (from `drug_specific_features.md`)

For each drug category expected in `all_drugs_filtered` (days −28 to −1, ≥10 use events in the window), this section records which of the three feature types to include and the justification.

**Feature types:**
- **days** — total days (any source) in the window with this substance
- **streak** — longest consecutive-day run with this substance
- **binary** — any use at all in the window (0/1)

Verify the exact list of surviving categories by running:
```r
all_drugs_filtered |> count(what_grouped) |> arrange(desc(n))
```
before finalising the analysis tibble.

### Alcohol

All three alcohol categories are renamed in `drug_map` to begin with "Alcohol" so they sort together alphabetically. No aggregation occurs — they remain distinct features.

#### Alcohol Heavy Amnt

| Feature | Include? | Justification |
|---|---|---|
| `alcohol_heavy_amnt_days` | ✅ Yes | Heavy drinking is a robust predictor of opioid relapse; distinct from moderate drinking |
| `alcohol_heavy_amnt_streak` | ✅ Yes | Heavy drinking streaks (bingeing) predict worse outcomes than isolated events |
| `alcohol_heavy_amnt_binary` | ✅ Yes | Any heavy drinking vs. none |

#### Alcohol Light Amnt

| Feature | Include? | Justification |
|---|---|---|
| `alcohol_light_amnt_days` | ✅ Yes | Include separately from heavy; captures overall drinking behaviour and restraint when used together |
| `alcohol_light_amnt_streak` | ❌ No | Light drinking streaks have low clinical significance for OUD prediction; omit to reduce noise |
| `alcohol_light_amnt_binary` | ✅ Yes | Any light drinking vs. abstinence |

#### Alcohol Missing Amnt

| Feature | Include? | Justification |
|---|---|---|
| `alcohol_missing_amnt_days` | ✅ Yes | Days where alcohol was detected but quantity was not recorded; captures data gaps and may carry its own under-reporting signal |
| `alcohol_missing_amnt_streak` | ❌ No | Streak has no interpretable meaning when quantity (heavy vs. light) is unknown. TODO: re-evaluate if usage patterns closely track heavy drinking |
| `alcohol_missing_amnt_binary` | ✅ Yes | Presence of unquantified alcohol records may correlate with under-reporting behaviour |

**Note:** The "restraint" feature (`alcohol_heavy_amnt_days − alcohol_light_amnt_days`) is a derived feature computed separately at tibble-building time (§6). Keep the raw counts as separate features too.

---

### Amphetamine

#### Amphetamine

| Feature | Include? | Justification |
|---|---|---|
| `amphetamine_days` | ✅ Yes | Stimulant co-use; distinct from methamphetamine in form, access pattern, and demographic associations |
| `amphetamine_streak` | ✅ Yes | Addictive with binge-pattern use; streak captures compulsive escalation |
| `amphetamine_binary` | ✅ Yes | Any vs. none is clinically meaningful |

---

### Analgesic

#### Analgesic (Gabapentin + Acetaminophen)

| Feature | Include? | Justification |
|---|---|---|
| `analgesic_days` | ✅ Yes | Prescription non-opioid pain/withdrawal management; functions as a medication engagement marker |
| `analgesic_streak` | ❌ No | Not established as a substance with compulsive-use streaks; streak in non-addictive Rx is semantically inverted (streak = compliance, not escalation) and would contaminate streak aggregations |
| `analgesic_binary` | ✅ Yes | Any gabapentin/acetaminophen use = baseline prescription engagement signal |

**Note:** Gabapentin has borderline misuse potential in OUD populations but is grouped here because its primary role in this dataset is as a legally prescribed non-opioid analgesic. Acetaminophen piggybacks on gabapentin's event count to survive the primary filter (5 + 23 = 28 combined events).

---

### Antiemetic

#### Antiemetic

| Feature | Include? | Justification |
|---|---|---|
| `antiemetic_days` | ✅ Yes | Proxy for withdrawal symptom severity and medication adherence |
| `antiemetic_streak` | ❌ No | Streak in non-addictive Rx is semantically inverted and would contaminate streak aggregations with a positive-valence signal |
| `antiemetic_binary` | ✅ Yes | "On antiemetics at baseline" indicates withdrawal management engagement |

---

### Benzodiazepine

#### Benzodiazepine

| Feature | Include? | Justification |
|---|---|---|
| `benzodiazepine_days` | ✅ Yes | Cross-addiction with opioids; one of the strongest predictors of overdose and relapse |
| `benzodiazepine_streak` | ✅ Yes | Physical dependence develops with sustained use; streak reflects severity |
| `benzodiazepine_binary` | ✅ Yes | Any benzo use at intake is a major clinical flag |

---

### Buprenorphine

#### Buprenorphine

| Feature | Include? | Justification |
|---|---|---|
| `buprenorphine_days` | ✅ Yes | Pre-study MOUD exposure; duration may predict readiness/tolerance |
| `buprenorphine_streak` | ❌ No | A long streak = consistent medication-taking (desirable), not compulsive use. Streak is semantically inverted here and would contaminate streak aggregations |
| `buprenorphine_binary` | ✅ Yes | Any prior buprenorphine use |

---

### Cannabinoids

#### Cannabinoids (THC + K2)

| Feature | Include? | Justification |
|---|---|---|
| `cannabinoids_days` | ✅ Yes | Cannabis use has a complex bidirectional relationship with opioid use; include to let the model find the direction |
| `cannabinoids_streak` | ✅ Yes | Cannabis use disorder is real; streak captures habitual use vs. occasional |
| `cannabinoids_binary` | ✅ Yes | Any use vs. none |

**Note:** THC and K2 (synthetic cannabinoid) are combined into one category via `drug_map`.

---

### Cocaine

#### Cocaine

| Feature | Include? | Justification |
|---|---|---|
| `cocaine_days` | ✅ Yes | Stimulant co-use strongly predicts opioid relapse (polydrug severity marker) |
| `cocaine_streak` | ✅ Yes | Binge-pattern cocaine use captured by streak; addictive |
| `cocaine_binary` | ✅ Yes | Any vs. no cocaine use is a meaningful severity split |

---

### Crack

#### Crack cocaine

| Feature | Include? | Justification |
|---|---|---|
| `crack_days` | ✅ Yes (if ≥10 events) | Distinct use pattern from powder cocaine |
| `crack_streak` | ✅ Yes (if ≥10 events) | Same rationale as cocaine |
| `crack_binary` | ✅ Yes (if ≥10 events) | |

**Note:** If both `cocaine_days` and `crack_days` survive the filter but `crack_days` is very sparse (10–30 events), consider whether a composite `stimulant_days = cocaine_days + crack_days` is more useful. Leave this decision to the ML pipeline unless counts are extremely low.

---

### Fentanyl

#### Fentanyl

| Feature | Include? | Justification |
|---|---|---|
| `fentanyl_days` | ✅ Yes | Increasingly prevalent; pharmacologically distinct from heroin (faster onset, higher potency) |
| `fentanyl_streak` | ✅ Yes | Same rationale as heroin |
| `fentanyl_binary` | ✅ Yes | Any fentanyl use is clinically meaningful (overdose risk, tolerance) |

---

### Heroin

#### Heroin

| Feature | Include? | Justification |
|---|---|---|
| `heroin_days` | ✅ Yes | Strongest single predictor of OUD relapse; graded dose-response with outcome |
| `heroin_streak` | ✅ Yes | Streak captures compulsive dependent use, not just occasional use |
| `heroin_binary` | ✅ Yes | Any heroin use vs. none is a clinically meaningful threshold; useful for linear models |

---

### MDMA/Hallucinogen

#### MDMA / Hallucinogen

| Feature | Include? | Justification |
|---|---|---|
| `mdma_hallucinogen_days` | ✅ Yes | Rare but present (passed primary filter); retained since secondary filter is not applied |
| `mdma_hallucinogen_streak` | ❌ No | Weak established link to OUD relapse; streak unlikely to carry additional signal |
| `mdma_hallucinogen_binary` | ✅ Yes | Any use flag retained for completeness |

---

### Methadone

#### Methadone

| Feature | Include? | Justification |
|---|---|---|
| `methadone_days` | ✅ Yes | Pre-study methadone exposure is relevant for CTN-0028 participants |
| `methadone_streak` | ❌ No | Same reasoning as buprenorphine — streak = compliance, semantically inverted |
| `methadone_binary` | ✅ Yes | Any prior methadone use |

---

### Methamphetamine

#### Methamphetamine

| Feature | Include? | Justification |
|---|---|---|
| `methamphetamine_days` | ✅ Yes | Increasingly common in OUD populations; high addiction potential |
| `methamphetamine_streak` | ✅ Yes | Binge/run use pattern; streak meaningful |
| `methamphetamine_binary` | ✅ Yes | Any vs. none split is clinically relevant |

---

### Muscle Relaxant

#### Muscle Relaxant

| Feature | Include? | Justification |
|---|---|---|
| `muscle_relaxant_days` | ✅ Yes | Passed primary filter; prescription engagement signal analogous to analgesic/antiemetic |
| `muscle_relaxant_streak` | ❌ No | Non-addictive Rx; streak semantically inverted and would contaminate streak aggregations |
| `muscle_relaxant_binary` | ✅ Yes | Any use flag; prescription adherence marker |

---

### Opioid

#### Opioid group (prescription: Codeine, Hydrocodone, Oxycodone, etc.)

| Feature | Include? | Justification |
|---|---|---|
| `opioid_days` | ✅ Yes | Prescription opioid misuse before study; distinct from heroin/fentanyl in route and access pattern |
| `opioid_streak` | ✅ Yes | Compulsive prescription opioid use is addictive; streak meaningful |
| `opioid_binary` | ✅ Yes | Any prescription opioid misuse vs. none |

---

### PCP

#### PCP (Phencyclidine)

| Feature | Include? | Justification |
|---|---|---|
| `pcp_days` | ✅ Yes | Frequency of use is a meaningful severity signal in a polydrug OUD context |
| `pcp_streak` | ✅ Yes | PCP has psychological dependence and habituation potential; without a withdrawal driver, long streaks reflect compulsive/habitual use rather than physical necessity — still clinically meaningful |
| `pcp_binary` | ✅ Yes | Any PCP use vs. none is a relevant behavioural marker in this population |

---

### Sedatives

#### Sedatives (Sedative-Hypnotic + Barbiturate)

| Feature | Include? | Justification |
|---|---|---|
| `sedatives_days` | ✅ Yes (if ≥10 events) | Non-benzo CNS depressants; hypnotics (Ambien etc.) remain common |
| `sedatives_streak` | ✅ Yes (if ≥10 events) | Physical dependence potential; streak meaningful if events present |
| `sedatives_binary` | ✅ Yes (if ≥10 events) | Any use flag |

---

### Suboxone

#### Suboxone

| Feature | Include? | Justification |
|---|---|---|
| `suboxone_days` | ✅ Yes | Treat as buprenorphine-equivalent; may combine with `buprenorphine_days` at modelling time |
| `suboxone_streak` | ❌ No | Same reasoning as buprenorphine |
| `suboxone_binary` | ✅ Yes | Combined with `buprenorphine_binary` for the "Took THEIR study drug" feature |

---

### Excluded categories

#### Antidepressant (Trazodone + Tricyclic)

Dropped by primary filter: 3 events across 3 people in the pre-randomisation window. Will be revisited as part of a future legal/prescription medicine engagement composite feature built from raw `all_drugs` labels before filtering.

#### Caffeine

Not included. No established relationship with OUD relapse; near-universal use creates near-zero variance.

---

## Summary table

| Drug | days | streak | binary |
|---|---|---|---|
| Alcohol Heavy Amnt | ✅ | ✅ | ✅ |
| Alcohol Light Amnt | ✅ | ❌ | ✅ |
| Alcohol Missing Amnt | ✅ | ❌ | ✅ |
| Amphetamine | ✅ | ✅ | ✅ |
| Analgesic (Gabapentin + Acetaminophen) | ✅ | ❌ | ✅ |
| Antiemetic | ✅ | ❌ | ✅ |
| Benzodiazepine | ✅ | ✅ | ✅ |
| Buprenorphine | ✅ | ❌ | ✅ |
| Cannabinoids (THC + K2) | ✅ | ✅ | ✅ |
| Cocaine | ✅ | ✅ | ✅ |
| Crack | ✅ | ✅ | ✅ |
| Fentanyl | ✅ | ✅ | ✅ |
| Heroin | ✅ | ✅ | ✅ |
| MDMA/Hallucinogen | ✅ | ❌ | ✅ |
| Methadone | ✅ | ❌ | ✅ |
| Methamphetamine | ✅ | ✅ | ✅ |
| Muscle Relaxant | ✅ | ❌ | ✅ |
| Opioid (rx group) | ✅ | ✅ | ✅ |
| PCP | ✅ | ✅ | ✅ |
| Sedatives | ✅ | ✅ | ✅ |
| Suboxone | ✅ | ❌ | ✅ |

✅ Include | ❌ Omit

---

# ── 6. Derived alcohol feature ────────────────────────────────────

## all_drugs

### Feature Proposal

Heavy drinking events - light drinking events.

Measures: restraint.

Include heavy drinking count and light drinking count as **separate features** in addition to (or instead of) their difference. The difference `heavy - light` collapses information: a person with 10 heavy and 5 light days is very different from one with 2 heavy and 0 light days, yet both produce the same difference of +5. The individual counts are independently predictive of relapse (Witkiewitz & Marlatt 2004), and tree-based ML models will recover the interaction automatically if both counts are present. Keeping them separate costs one column and loses nothing.

---

# ── 7. Polydrug & breadth features ───────────────────────────────

## all_drugs

### Feature proposal

**Polydrug use days**: the number of days in the -28 to 0 window on which the participant used 2 or more distinct drug categories simultaneously (as recorded in `all_drugs_filtered`). Polydrug use at baseline is a consistent predictor of worse OUD treatment outcomes (Shand et al. 2011; Kidorf et al. 2004) and captures a severity dimension not captured by any single-drug count.

### Dataset Notes

#### Feature proposal

**Total unique drug categories used** in the -28 to 0 window. A scalar measure of the breadth of substance use, independently of individual drug counts. Higher drug-use breadth predicts worse treatment retention (Stitzer & Sigmon 2006).

---

# ── 8. Lie count ─────────────────────────────────────────────────

## all_drugs

### Feature proposal

> **⚠️ DROPPED — `lie_count` / `lie_rate`**: For nearly all participants there were no UDS or UDSAB screenings in the pre-trial window. For those with screenings, self-report was more or less consistent with UDS results, up to drug category grouping name differences. This resulted in `lie_count` and `lie_rate` being `NA` or `0` for the vast majority of participants. These features carry no predictive signal and have been excluded from the analysis tibble.

~~Lie count. The number of days (not drugs, etc. days.) each person's drug results self reported from tlfb did not match their reports from UDS or UDSAB. If they over-reported something in tlfb that was not in uds/udsab, do not up the count. Only up the days count for under-reporting.~~

~~Also compute a **normalized lie rate**: `lie_count / total_drug_positive_days` (where `total_drug_positive_days` is the number of days UDS or UDSAB detected any substance). The raw count is confounded by baseline drug use intensity — a person who used drugs 20 days has more opportunities to under-report than one who used 5 days. The rate controls for this. Include both the raw count and the rate as separate features. If `total_drug_positive_days = 0`, set rate to 0.~~

---

# ── 9. ASI features ───────────────────────────────────────────────

## asi

### Dataset Notes

```{r asi-overview}
glimpse(asi)
```

#### Feature proposal

**IV drug use binary (ASI)**: Whether the participant has ever used drugs intravenously, as recorded in the `asi` dataset. IV drug use is associated with more severe OUD, higher HIV risk, and worse treatment outcomes (Roux et al. 2014; Gossop et al. 2002). This is distinct from `rbs_iv` which captures recent IV behavior — the ASI item captures lifetime history.

#### Feature proposal

**ASI composite scores**: If the dataset contains Addiction Severity Index composite subscores (drug, alcohol, family/social, legal, psychiatric, employment, medical), include all of them. The ASI drug composite is among the strongest validated predictors of OUD treatment outcome (McLellan et al. 1992). The family/social composite captures social support deficits. The psychiatric composite captures comorbidity load. All seven composites are defensible inclusions under current clinical consensus.

---

# ── 10. Demographics features ─────────────────────────────────────

## demographics

### who

### age

```{r age-plot}
demographics |>
  left_join(everybody, by = "who") |>
  ggplot(aes(x = age)) +
  geom_histogram() +
  facet_wrap(~project)
```

In addition to raw `age`, consider including **age at first opioid use** if available in the data (check `asi` or `rbs`). Earlier age of first use is a robust predictor of worse outcomes (Anthony & Petronis 1995). If not available, include age as a continuous feature, but also consider whether the ML pipeline will benefit from an age-squared term to capture the non-linear (inverted-U) relationship between age and relapse risk. Tree-based models handle this automatically; linear models will not.

### is_hispanic

Notes: Study 30 has about 5% hispanic while the others have ~17%

```{r is-hispanic-summary}
demographics |>
  left_join(everybody, by = "who") |>
  select(who, project, is_hispanic) |>
  mutate(is_hispanic = tolower(is_hispanic)) |>
  mutate(
    hispanic_binary = case_match(
      is_hispanic,
      "yes" ~ 1,
      "no"  ~ 0,
      .default = 0
    )
  ) |>
  group_by(project) |>
  summarize(
    hispanic_count     = sum(hispanic_binary),
    total_participants = n(),
    prop_hispanic      = hispanic_count / total_participants
  )
```

### race

```{r race-plot}
race_by_project <- demographics |>
  left_join(everybody, by = "who") |>
  select(who, project, race) |>
  group_by(project, race) |>
  summarize(count = n(), .groups = "drop")

ggplot(race_by_project, aes(x = "", y = count, fill = race)) +
  geom_bar(stat = "identity", width = 1) +
  labs(title = "Race Distribution by Project") +
  facet_wrap(~project, scales = "free") +
  coord_polar(theta = "y") +
  theme_void()
```

### job

```{r job-plot}
demographics |>
  left_join(everybody, by = "who") |>
  select(who, project, job) |>
  group_by(project, job) |>
  summarize(count = n(), .groups = "drop") |>
  ggplot(aes(x = job, y = count, fill = project)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_viridis_d() +
  labs(title = "Job Status Distribution by Project") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### is_living_stable

```{r living-stable-plot}
demographics |>
  left_join(everybody, by = "who") |>
  select(who, project, is_living_stable) |>
  group_by(project, is_living_stable) |>
  summarize(count = n(), .groups = "drop") |>
  ggplot(aes(x = is_living_stable, y = count, fill = project)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_viridis_d() +
  labs(title = "Housing Stability Distribution by Project")
```

#### Feature proposal

Include, but be careful by this causing weird behavior with the distribution.

Confirmatory Factor: Lack of support

### education

```{r education-plot}
demographics |>
  left_join(everybody, by = "who") |>
  select(who, project, education) |>
  group_by(project, education) |>
  summarize(count = n(), .groups = "drop") |>
  ggplot(aes(x = education, y = count, fill = project)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_viridis_d() +
  labs(title = "Education Distribution by Project") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### marital

```{r marital-plot}
demographics |>
  left_join(everybody, by = "who") |>
  select(who, project, marital) |>
  group_by(project, marital) |>
  summarize(count = n(), .groups = "drop") |>
  ggplot(aes(x = marital, y = count, fill = project)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_viridis_d() +
  labs(title = "Marital Status Distribution by Project")
```

#### Feature proposal

Include marital. Those who are married have a built in reminder to take their medication.

Confirmatory factor: Lack of support

### is_male

```{r is-male-plot}
demographics |>
  left_join(everybody, by = "who") |>
  select(who, project, is_male) |>
  group_by(project, is_male) |>
  summarize(count = n(), .groups = "drop") |>
  ggplot(aes(x = is_male, y = count, fill = project)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_viridis_d() +
  labs(title = "Sex Distribution by Project")
```

### Dataset Notes

I think each of the above should be features. Since these are categorical variables, think about how to structure them.

---

# ── 11. Fagerstrom ────────────────────────────────────────────────

## fagerstrom

### Dataset Notes

```{r fagerstrom-overview}
glimpse(fagerstrom)
```

#### Feature proposal

**Fagerstrom total score** (or its raw components). The Fagerstrom Test for Nicotine Dependence (FTND) is a validated 6-item scale. Nicotine dependence severity is one of the most consistently reported predictors of OUD treatment failure — smokers have significantly lower abstinence rates from opioids at 12 months (Weinberger et al. 2018; Tsoh et al. 2011). Mechanistically, shared dopaminergic pathways mean nicotine craving and opioid craving are not independent. Include as a continuous score.

---

# ── 12. Pain: baseline scalar ─────────────────────────────────────

## pain

```{r pain-overview}
glimpse(pain)
```

### Dataset Notes

`pain$when` ranges from −1 to 955. Pain is a **cross-sectional baseline measure** — one row per participant collected at the intake visit (SF-36 for CTN-0027/0030, EuroQoL for CTN-0051). Not a daily longitudinal variable; window aggregates are meaningless. Longitudinal daily self-medication features have moved to §22/§23 using `withdrawal` (COWS/SOWS) as the correct daily measure.

#### Feature proposal

**`pain_score`** — ordinal integer: 0 = No Pain, 1 = Very mild to Moderate Pain, 2 = Severe Pain, NA = Missing. Scalar baseline pain severity at intake.

Self-reported baseline pain is independently associated with OUD treatment dropout and relapse (Rhodin et al. 2006; Dunn et al. 2014). Used as a scalar in psychiatric interaction terms (§25).

---

# ── 13. Psychiatric features ──────────────────────────────────────

## psychiatric

### Dataset Notes

```{r psychiatric-overview}
glimpse(psychiatric)
```

#### Feature proposal

The `psychiatric` dataset provides 18 features. All Yes/No flags are encoded 0/1; NA = not assessed.

**Medical history diagnosis flags** (from clinical interview):

| Feature | Condition | Notes |
|---|---|---|
| `has_schizophrenia` | Schizophrenia | |
| `has_major_dep` | Major depression | Strongest psychiatric predictor of OUD relapse (Nunes & Levin 2004) |
| `has_bipolar` | Bipolar disorder | |
| `has_anx_pan` | Anxiety / panic disorder | Predicts worse XR-NTX outcomes (Lee et al. 2018) |
| `has_brain_damage` | Brain damage | |
| `has_epilepsy` | Epilepsy | |

**DSM substance use disorder diagnoses:**

| Feature | Condition |
|---|---|
| `has_opiates_dx` | Opioid use disorder (DSM-IV/5) |
| `has_alcol_dx` | Alcohol use disorder |
| `has_amphetamines_dx` | Amphetamine-type SUD |
| `has_cannabis_dx` | Cannabis use disorder |
| `has_cocaine_dx` | Cocaine use disorder |
| `has_sedatives_dx` | Sedative/hypnotic/anxiolytic SUD |

**ASI self-report items** (Yes → 1, No → 0, "Not answered" / "Missing" → NA):

| Feature | ASI Item |
|---|---|
| `depression_binary` | P4: Serious depression / sadness / hopelessness / difficulty functioning |
| `anxiety_binary` | P5: Serious anxiety / tension / inability to relax |
| `schizophrenia_binary` | P6: Hallucinations |

**Derived:**

- **`psych_comorbidity_count`**: sum of the 6 medical history flags. Captures total psychiatric burden; higher load predicts worse OUD treatment outcomes (Watkins et al. 2011).
- **`substance_dx_count`**: sum of the 5 non-opioid SUD diagnoses (`has_opiates_dx` excluded — near-universal in this sample). Captures polysubstance disorder burden.
- **`dep_or_anx_binary`**: 1 if `depression_binary == 1` OR `anxiety_binary == 1`. The internalizing comorbidity cluster is the most replicated psychiatric predictor of OUD relapse; co-occurring depression roughly halves abstinence rates in MOUD trials (Nunes & Levin 2004).

Cross-dataset interactions (psychiatric × withdrawal, psychiatric × benzodiazepine) are in §25.

---

# ── 14. Quality of life ───────────────────────────────────────────

## qol

### Dataset Notes

```{r qol-overview}
glimpse(qol)
```

**Note: CTN-0051 only.** The `qol` dataset contains 657 rows, all from project 51. The single usable feature is `is_homeless` — a Yes/No indicator of current homelessness at baseline. The SF-36/EQ-5D composite scores originally proposed here do not exist in this dataset.

#### Feature proposal

**`is_homeless` binary** (0/1 integer). Objective housing status at baseline. Homelessness is among the strongest predictors of OUD treatment dropout and relapse across MOUD modalities — it captures housing instability, lack of safe space to store/take medication, and social environment risk in a single binary flag (Tsemberis et al. 2004; Palepu et al. 2006).

> **⚠️ Cross-dataset note: `is_living_stable` × `is_homeless` discordance (see §26)**
>
> `is_living_stable` (demographics, all studies) and `is_homeless` (qol, CTN-0051 only) are related but distinct constructs. An inner join reveals **128 participants coded (Yes, Yes)** — stable living AND homeless simultaneously. All 128 are CTN-0051; this reflects shelter residents who perceive their shelter as "stable." The two variables measure different things: **subjective perceived stability** vs. **objective permanent housing status**. Including both as raw binaries masks this and creates near-collinearity for some patterns. See §26 for the `housing_stability_ord` ordered factor that combines them explicitly.

---

# ── 15. Treatment arm ─────────────────────────────────────────────

## randomization

Note: make sure to include first randomization only.

#### Feature proposal

**Treatment arm** (what medication the participant was randomized to: methadone, buprenorphine-naloxone, or XR-naltrexone). This is arguably the most important covariate in the entire analysis. The X:BOT trial (CTN-0051; Lee et al. 2018) found that XR-NTX had significantly lower retention than buprenorphine-naloxone, primarily due to the induction barrier. If treatment arm is not included as a feature, the model will confound medication effects with patient characteristics. Include as a categorical variable (one-hot or ordinal depending on model).

Note: the `treatment` column in `first_rand` already uniquely identifies the study — the 6 arm labels are distinct across all three projects (27, 30, 51). No join to `everybody` is needed.

---

# ── 16. RBS features ─────────────────────────────────────────────

## rbs

### Dataset Notes

```{r rbs-overview}
glimpse(rbs)
```

**Note: rbs dropped from main feature set.** The dataset contains only 5 drug categories (cocaine, heroin, speedball, opioid, speed) with bucketed day counts (0/4/30) — a coarser, same-source-type duplicate of the −28 to 0 drug event data already captured at higher resolution in `drug_feats`. The only unique signal extracted is the heroin denial flag (`heroin_rbs_alldr_incons`), computed as a cross-dataset feature in §24.

#### Feature proposal

**Social network drug use**: The number or proportion of people in the participant's social network who use drugs (if this item is in the RBS). Social network drug use is one of the strongest documented predictors of OUD relapse — having drug-using peers roughly doubles relapse odds at 12 months (Havassy et al. 1991; Jason et al. 2010). This captures environmental exposure that no within-person drug history variable can capture.

#### Feature proposal

**Number of recent sexual partners**: From the RBS. High partner count is a marker of HIV risk behavior and a proxy for social instability and sensation-seeking, both of which predict OUD treatment dropout (Nunes et al. 2010). Include as a continuous or binned count.

> <h2 style="color: #c0392b; background: #fdecea; padding: 10px; border-left: 6px solid #c0392b; font-size: 1.4em;">⚠️ QUESTION FOR DR. BALISE: Why are social network drug use and number of recent sexual partners missing from RBS? These are standard RBS items and would be valuable features. See Email.</h2>

---

# ── 17. RBS_IV: recent IV drug use ───────────────────────────────

## rbs_iv

### Dataset Notes

```{r rbs-iv-overview}
glimpse(rbs_iv)
```

#### Feature proposal

All 9 columns included as features. IV drug use at intake is associated with more severe OUD, greater withdrawal intensity, higher overdose risk, and substantially worse treatment retention in buprenorphine and XR-NTX arms (Roux et al. 2014; Gossop et al. 2002). This is distinct from the ASI lifetime IV use indicator — the recent/current measure is more clinically proximal.

| Feature | Source column | Notes |
|---|---|---|
| `iv_days_total` | `days` | Max days of IV use across all drugs; definitionally = max of per-drug columns — expect high correlation |
| `iv_events` | `max` | Total injection events (not days) |
| `iv_amount_dominant` | `amount` | Total exposure for most-used IV drug |
| `iv_shared_needles` | `shared` | Binary; needle sharing = HIV risk + social instability proxy |
| `cocaine_inject_days` | `cocaine_inject_days` | NA → 0 (did not inject cocaine) |
| `heroin_inject_days` | `heroin_inject_days` | NA → 0; will correlate with speedball |
| `speedball_inject_days` | `speedball_inject_days` | NA → 0; contains heroin, expect heroin correlation |
| `opioid_inject_days` | `opioid_inject_days` | NA → 0 |
| `speed_inject_days` | `speed_inject_days` | NA → 0 |

`iv_days_total`, `iv_events`, `iv_amount_dominant`: NA left as-is (genuinely missing aggregate). `iv_shared_needles`: NA left as-is (missing response).

---

# ── 18. Study site ────────────────────────────────────────────────

## site_masked

### Dataset Notes

#### Feature proposal

**Study site (masked)**: Include `site_masked` as a covariate. Multi-site RCTs routinely show substantial site effects on retention and relapse rates, driven by differences in clinic culture, staff support, and local drug market conditions. Omitting site risks attributing site-level variance to individual features. Given the masked coding, this can be included directly as a factor with dummy/one-hot encoding.

---

# ── 19. Withdrawal pre/post ───────────────────────────────────────

## withdrawal_pre_post

### Dataset Notes

```{r withdrawal-pre-post-overview}
glimpse(withdrawal_pre_post)
```

`withdrawal_pre_post` contains no negative `when` values — all observations are at day 0 or later (post-randomization). The `what` column marks rows as `"pre"` (before first MOUD dose) or `"post"` (after first MOUD dose). "Pre" and "post" refer to **induction**, not study enrollment. The dataset therefore covers the post-enrollment, pre-/peri-induction window.

> <h2 style="color: #c0392b; background: #fdecea; padding: 10px; border-left: 6px solid #c0392b; font-size: 1.4em;">⚠️ QUESTION FOR DR. BALISE: Can we use the pre-induction rows (what == "pre", when ≥ 0) as baseline features? They are pre-treatment but post-enrollment — is that acceptable for a baseline predictor?</h2>

> <div style="color: #ffffff; background: #c0392b; padding: 14px; border: 4px solid #922b21; font-size: 1.5em; font-weight: bold; margin-top: 12px;">🚨 CRITICAL — NOTE TO NAT: WINDOW ANCHOR MAY BE WRONG 🚨<br><br>Every longitudinal feature in this analysis filters <code>when &gt;= -28 &amp; when &lt; 0</code>, anchored to <strong>randomization (day 0)</strong>. But if induction happens days after randomization, data in the [0, induction_day) range is valid pre-treatment baseline that we are currently discarding entirely.<br><br>If the window should instead be anchored to <strong>induction date</strong>, every single drug feature (§5), pain feature, rbs_iv feature, and every other longitudinal section is filtered incorrectly. This is the highest-stakes unresolved methodological question in the analysis.<br><br>We do NOT have an induction date column. It may be inferable from <code>withdrawal_pre_post</code> (the minimum positive <code>when</code> where <code>what == "pre"</code> per participant), but this is unverified. Confirm with Dr. Balise before modelling.</div>

#### What `screening_date` tells us about the window

The `screening_date` dataset contains a `day_zero` column described as "best guess at the end of TLFB" — i.e., the day (relative to randomization) when each participant's Timeline Followback assessment was actually collected. This is the per-person anchor for the `all_drugs` window, and it is **not always 0**.

```{r screening-date-overview}
glimpse(screening_date)
```

```{r day-zero-distribution}
hist(
  screening_date$day_zero,
  breaks = 50,
  main   = "Distribution of day_zero (TLFB anchor relative to randomization)",
  xlab   = "day_zero (days)",
  ylab   = "Participants"
)
```

The distribution spikes sharply at 0 or −1, meaning the majority of participants had their TLFB collected at or immediately before randomization — so our fixed `when >= -28 & when < 0` filter is approximately correct for most people. However, there is a meaningful left tail reaching −28 and a right tail out to +20. Participants far from zero have their entire TLFB window shifted: participant 14 (`day_zero = 27`) has essentially none of their drug records captured by our current filter.

**Bottom line:** the error is real but probably acceptable for the majority of the sample. Whether to implement per-person `day_zero`-anchored windows is a methodological call for Dr. Balise — do not change the filter without that conversation.

#### Feature proposal

**Pre-induction withdrawal severity score** (baseline COWS or SOWS score before first dose). Baseline withdrawal severity is a predictor of buprenorphine induction success and XR-NTX induction completion. Patients with higher pre-induction COWS are more likely to fail induction onto XR-NTX (Lee et al. 2018; Sigmon et al. 2012), creating a selection effect that is critical to model. Include as a continuous score.

#### Feature proposal

**Delta withdrawal score** (post-induction score minus pre-induction score, or equivalently: how much withdrawal improved after the first dose). Larger improvement ("responders") are more likely to remain in treatment. This captures individual pharmacokinetic responsiveness beyond baseline severity alone. Derivable from `withdrawal_pre_post`.

---

# ── 20. Database Notes: Took THEIR / DIFFERENT study drug ─────────

## all_drugs × (randomization × everybody)

### Feature Proposal

Took THEIR study drug before study

### Feature Proposal

Took DIFFERENT study drug before study (e.g. their study put them on Bup but they had attempted to get clean and failed pre study with xr-ntx)

Notes for above 2 features:

What are the study drugs?

- CTN-0028: Methadone vs Buprenorphine-Naloxone
- CTN-0030: Only Buprenorphine-Naloxone (Suboxone)
- CTN-0051: XR Naltrexone vs Buprenorphine-Naloxone

- You need to not only get what study but also what drug they were randomized to
- In the `all_drugs_filtered` table I am using, Buprenorphine and Suboxone are separate. If someone is randomized to Suboxone but took Buprenorphine prior, count this as having taken THEIR study drug. They are substantially similar.
- We do not need "Took no study drug before study" because it should be collinear with the above two and probably would break stuff

---

# ── 21. Withdrawal: baseline scalar ──────────────────────────────

## withdrawal_pre_post

### Dataset Notes

```{r withdrawal-overview}
glimpse(withdrawal_pre_post)
```

`withdrawal_pre_post |> filter(what == "pre")` contains exactly one pre-induction COWS/SOWS observation per participant — the baseline severity score before the first MOUD dose. CTN-0027/0030 use COWS; CTN-0051 uses SOWS. Factor levels: 0=None, 1=mild, 2=moderate, 3=severe.

This is a scalar baseline measure, not a daily longitudinal variable. Treated identically to `pain_score` (§12).

#### Feature proposal

- **`withdrawal_pre_score`**: Pre-induction COWS/SOWS severity (0=None, 1=mild, 2=moderate, 3=severe). One row per person from `withdrawal_pre_post` "pre" records.

> **NOTE: windowed longitudinal aggregates dropped.** `withdrawal_pre_post` contains exactly 1 "pre" obs per person; windowed aggregates (mean, max, pct days severe/any) collapse to this scalar and add no information. Code retained in §21 as comments for reference.

---

# ── 22. Database Notes: Withdrawal × hard drug interaction ────────

## Database Notes

### Feature proposal: withdrawal x hard drug

Hypothesis: participants who use more HARD drugs on the days they are experiencing more withdrawal severity are MORE likely to relapse. Pain was originally proposed here, but pain is a cross-sectional scalar (one row per person). `withdrawal` (daily COWS/SOWS) is the correct daily longitudinal measure.

**Hard drug composite**: Heroin, Fentanyl, Cocaine, Crack, Methamphetamine, Amphetamine, Opioid (prescription opioids via `drug_map`). Opioid included because prescription opioid misuse on high-withdrawal days is the core self-medication pathway. Does NOT include MOUD (Buprenorphine/Suboxone/Methadone are separate pass-through categories).

**Feature 1 — `withdrawal_harddrug_corr`**: Person-level Pearson correlation between daily withdrawal score (0–3) and daily hard-drug-use binary across the pre-study window. If 0 hard drug days: coded as 0. If all days had hard drug use (zero variance): coded as NA.

**Feature 2 — `withdrawal_highrisk_harddrug_rate`**: Among days where the person's withdrawal score exceeds their own median for the window, the proportion of those days that also had hard drug use. Directly operationalises "self-medicating high-withdrawal days."

> ~~**NOTE: Implementation deferred.**~~ **DROPPED.** Correlation features require ≥2 longitudinal observations per person per window. `withdrawal_pre_post` contains exactly 1 "pre" observation per person. Code retained in §22 as comments for reference.

Confirmatory factor: Self-medication tendency

---

# ── 23. Database Notes: Withdrawal × soft drug interaction ────────

## Database Notes

### Feature proposal: withdrawal x soft drug

Hypothesis: participants who use SOFT drugs (cannabis, light alcohol) on high-withdrawal days are LESS likely to relapse — substituting non-opioids for opioids when withdrawal peaks. Pain was originally proposed here, but pain is a cross-sectional scalar.

**Soft drug composite**: Cannabinoids, Alcohol Light Amnt. Caffeine removed — not present in `all_drugs_grouped` within the window. Alcohol Heavy Amnt excluded (not a "soft" use pattern).

Apply the same two-feature operationalisation as withdrawal × hard drugs above:

- **`withdrawal_softdrug_corr`**: Pearson correlation between daily withdrawal score and daily soft-drug binary across the pre-study window. Expect negative (protective direction).
- **`withdrawal_highrisk_softdrug_rate`**: Proportion of above-personal-median withdrawal days on which any soft drug was used. Protective-substitution indicated by higher rate being associated with lower relapse risk.

> ~~**NOTE: Implementation deferred.**~~ **DROPPED.** Same reason as §22 — single observation per person makes correlation undefined. Code retained in §23 as comments.

Confirmatory factor: Self-medication tendency

---

# ── 24. Database Notes: Heroin denial — all_drugs × rbs ─────────────

## all_drugs × rbs

### Dataset Notes

`heroin_days` (from `drug_feats`, §5) counts observed heroin events in the −28 to 0 window from `all_drugs`. The `rbs` dataset (§16) asks participants to self-report how many days they used heroin (and speedball, which contains heroin) in the past 30 days. These two instruments measure the same construct from different angles.

#### Feature proposal

**`heroin_rbs_alldr_incons`** (0/1 integer). Flags participants whose `all_drugs` records show ≥1 heroin day in the window but who denied heroin use in both the heroin and speedball rbs rows. Candidate proxy for minimization of the primary drug of abuse — a known predictor of treatment non-engagement.

Heroin is the best-recorded drug in `all_drugs` (81k events), making false positives from missed records unlikely. Not confounded by MOUD (Buprenorphine/Methadone are distinct drug names) and not an inference artifact. ~5.9% prevalence in the sample (~146 of 2,490 participants).

> **Note on discrepancy vs Appendix A.8**: This section's implementation differs from Appendix A by ~6 participants, likely people who used speedball (heroin+cocaine) and did not separately report heroin use in the heroin rbs row.

Confirmatory factor: Denial/minimization of primary drug of abuse

---

# ── 25. Database Notes: Psychiatric × withdrawal and psychiatric × drug interactions ──

## psychiatric × wdl_main_feats × drug_feats

#### Feature proposal

Three interaction features computed in `psych_cross_feats` from §13 (`psych_feats`), §21 (`wdl_main_feats`), and §5 (`drug_feats`). Pain scalar interactions removed — `pain_score × has_major_dep` etc. are near-collinear with the psychiatric binary itself and add no information beyond the main effect.

| Feature | Formula | Clinical rationale |
|---|---|---|
| `has_major_dep_x_withdrawal_pre_score` | `has_major_dep × withdrawal_pre_score` | Depression × baseline withdrawal severity |
| `has_anx_pan_x_withdrawal_pre_score` | `has_anx_pan × withdrawal_pre_score` | Anxiety × baseline withdrawal severity |
| `has_anx_pan_x_benzodiazepine_days` | `has_anx_pan × benzodiazepine_days` | Anxiety + benzo co-use = elevated pharmacological overdose risk |

`withdrawal_mean`/`withdrawal_max` collapsed to `withdrawal_pre_score` (single baseline observation per person — see §21 note). Two interaction columns instead of four.

**Depression × withdrawal**: Depression lowers pain threshold, depletes dopaminergic reward circuitry, and is bidirectionally reinforced by withdrawal severity. In OUD populations, severe withdrawal in a depressed patient predicts opioid escalation and treatment dropout (Sheng et al. 2017; Rhodin et al. 2006).

**Anxiety × withdrawal**: Anxiety amplifies withdrawal-driven craving through hyperarousal and interoceptive sensitivity. A participant with anxiety disorder who enters treatment with high withdrawal severity faces a compounded craving load.

**Anxiety × benzodiazepine**: Signal is pharmacological risk. Benzodiazepine + opioid co-use synergistically depresses CNS respiratory drive, dramatically elevating overdose risk (FDA black box warning 2016; Park et al. 2015). `benzodiazepine_days` NA imputed to 0 before multiplying (no use = 0 days).

Confirmatory factors: Withdrawal-depression cycle; anxiety-mediated craving amplification; pharmacological overdose risk profile.

---

# ── 26. Database Notes: Housing stability (cross-dataset) ──────────

## demographics × qol

### Dataset Notes

`is_living_stable` (demographics, all studies) encodes subjective perceived housing stability. `is_homeless` (qol, CTN-0051 only) encodes objective homelessness status. These measure related but distinct constructs; the 128 (Yes,Yes) cases are shelter residents — objectively homeless but with stable shelter arrangements.

#### Feature proposal

**`housing_stability_ord`** — 3-level ordered factor, worst → best:

| Level | Cases `(is_living_stable, is_homeless)` | Construct |
|---|---|---|
| `unstable_homeless` (worst) | (No,Yes), (NA,Yes) | Confirmed homeless and not shelter-stable |
| `mid_stability` | (Yes,Yes), (No,No), (No,NA) | Shelter-stable; housed but precarious; unstable + homeless unknown |
| `stable_housed` (best) | (Yes,No), (NA,No), (Yes,NA) | Clearly stable; not-homeless drives; stable implies housed (non-CTN-0051) |
| NA | (NA,NA) | Both unknown |

**Rationale**: A shelter resident (Yes,Yes) is meaningfully less precarious than someone with no stable shelter (No/NA, Yes). Housed-but-unstable (No,No) — e.g., abusive household, couch-surfing — is worse than clearly stable (Yes,No) but better than literal homelessness. For non-CTN-0051 participants, `is_homeless` is NA; level assigned purely from `is_living_stable`. The ordinal captures this gradient rather than losing information via two correlated binaries.

Confirmatory factor: Social determinants of health; housing instability as relapse risk

---

# ── 27. Database Notes: Withdrawal trajectory ─────────────────────

## withdrawal

## Database Notes

### Feature proposal: withdrawal trajectory x time

**Slope of COWS/SOWS withdrawal scores over the −28 to 0 window** (from `withdrawal`). If a participant's withdrawal symptoms are *increasing* in the days before study entry (escalating dependence), this predicts worse outcomes compared to stable or decreasing withdrawal. Estimated per-person via `lm(score ~ when)` within the window (minimum 2 observations required). Increasing withdrawal slope has been associated with higher opioid craving and lower treatment readiness (O'Brien & McLellan 1996).

> **DROPPED.** Slope estimation via `lm(score ~ when)` requires ≥2 observations per person. `withdrawal_pre_post` contains exactly 1 "pre" observation per person. Code retained in §27 as comments for reference.

Confirmatory factor: Escalating dependence severity

---

# ── 28. Database Notes: Medication adherence composite ────────────

## Database Notes

### Feature proposal: medication adherence history (cross-dataset)

**Proposed prescribed-medication composite features** (derived from `all_drugs_filtered`, restricted to the categories below):

| Feature | Definition | Rationale |
|---|---|---|
| `rx_any_binary` | 1 if participant has ≥1 day of any prescribed non-MOUD medication in −28 to −1 | Binary adherence flag; "responsibility towards health" construct |
| `rx_days` | Total days in window with any prescribed non-MOUD medication recorded | Graded adherence signal |
| `rx_categories` | Count of distinct prescribed medication categories used | Breadth of medical engagement |

**Categories that contribute to the composite** (excludes MOUD and legal non-Rx substances):

- `Antidepressant` — psychiatric medication adherence
- `Analgesic` (Gabapentin + Acetaminophen) — pain/prescription management
- `Muscle Relaxant` — prescribed non-addictive medication
- `Antiemetic` — anti-nausea prescription (e.g. ondansetron)
- `Sedatives` / `Benzodiazepine` — prescribed CNS depressants *(note: these ARE addictive; consider excluding from the "non-addictive" composite and keeping as separate features)*

**Implementation note:** Use `all_drugs_filtered` filtered to these categories, then aggregate per person over the −28 to −1 window. Do not include MOUD (Buprenorphine, Suboxone, Methadone) in this composite — those have their own dedicated features and carry a different signal.

Confirmatory factor: Responsibility towards health

---

# ── 29. Final assembly ────────────────────────────────────────────

<!-- No source-document content maps here. Technical R implementation of the full left_join sequence and NA-fill logic. See building_analysis.R §29. -->

---

# Appendix: Flagged — Unmapped Content

The following sections from `Features_To_Include_Accepted_Suggestions.Rmd` have no corresponding `building_analysis.R` section and could not be placed in the ordered structure above.

> ⚠️ **FLAGGED** — `# meta_study_length`: No `building_analysis.R` section maps here. Original note: "Inclined to not include any features from here. If I do, 'treatment' and 'does_study_taper' will be the features. See comment on 'project'."

> ⚠️ **FLAGGED** — `# tlfb`: No `building_analysis.R` section maps here. Original note: "The following features have been removed because all_drugs has been filtered to have the level of granularity I desire."

> ⚠️ **FLAGGED** — `# first_survey`: Empty stub in original Rmd (header only, no content). No features. No `building_analysis.R` section.

> ⚠️ **FLAGGED** — `# meta_substance_groups_uds`: Empty stub in original Rmd (header only, no content). No features. No `building_analysis.R` section.

> ⚠️ **FLAGGED** — `# detox`: Empty stub in original Rmd (header only, no content). No features. No `building_analysis.R` section.

> ⚠️ **FLAGGED** — `# screening_date`: Empty stub in original Rmd (header only, no content). No features. No `building_analysis.R` section.

> ⚠️ **FLAGGED** — `# sex`: Empty stub in original Rmd (header only, no content). No features. No `building_analysis.R` section.

> ⚠️ **FLAGGED** — `# treatment`: Empty stub in original Rmd (header only, no content). No features. No `building_analysis.R` section. (Distinct from the treatment arm feature in §15, which comes from `randomization`.)

> ⚠️ **FLAGGED** — `# uds`: Empty stub in original Rmd (header only, no content). No standalone feature section. (UDS and UDSAB data were used in the §8 lie count computation, which was subsequently dropped — see §8 note.)

> ⚠️ **FLAGGED** — `# uds_temp`: Empty stub in original Rmd (header only, no content). No features. No `building_analysis.R` section.

> ⚠️ **FLAGGED** — `# visit`: Empty stub in original Rmd (header only, no content). No features. No `building_analysis.R` section.

> ⚠️ **FLAGGED** — `# withdrawal` (stub): Empty stub in original Rmd (header only, no content beneath it). The withdrawal trajectory feature proposal originates in the Database Notes section of the original Rmd and is placed in §27 above via that route. The `# withdrawal` header itself had no content.
