---
title: "Column Grouping Audit — analysis_tibble"
author: "Nat Oliven"
date: "`r Sys.Date()`"
output: html_document
---

# Setup

Run from project root (`ctn0094.oliven.Rproj`). Sources `building_analysis.R` if
`analysis_tibble` isn't already in the environment.

```{r setup, message=FALSE, warning=FALSE}
library(here)  # finds project root regardless of where this Rmd lives
if (!exists("analysis_tibble")) {
  library(conflicted)
  suppressPackageStartupMessages(library(tidymodels))
  tidymodels_prefer()
  suppressPackageStartupMessages(library(tidyverse))
  library(public.ctn0094data)
  conflicts_prefer(dplyr::filter)
  options(dplyr.summarise.inform = FALSE)
  source(here("vignettes/building_analysis.R"))
}
library(knitr)
```

---

# Batch Audit — All 115 Columns

**Primary output for Claude.** Saves `inst/scratch/col_audit_output.csv`. Run this
document once; Claude reads that file — no sequential back-and-forth needed.

```{r batch-audit}
col_audit <- analysis_tibble |>
  map_dfr(\(x) tibble(
    r_type     = paste(class(x), collapse = "/"),
    n_na       = sum(is.na(x)),
    pct_na     = round(mean(is.na(x)) * 100, 1),
    n_distinct = n_distinct(x, na.rm = TRUE),
    min_val    = if (is.numeric(x)) min(x, na.rm = TRUE) else NA_real_,
    max_val    = if (is.numeric(x)) max(x, na.rm = TRUE) else NA_real_,
    levels_str = if (is.factor(x)) paste(levels(x), collapse = " | ") else NA_character_
  ), .id = "column_name")

write_csv(col_audit, here("inst/scratch/col_audit_output.csv"))
col_audit |> kable(digits = 1)
```

---

# Spot Checks for Uncertain Columns

## Potential Zero-Variance Columns

Glimpse shows all-same values for several columns. ZV columns are useless to the
model and will be caught by `step_zv()`, but good to know ahead of time.

```{r zv-check}
zv_candidates <- c("heroin_binary", "heroin_days", "is_smoker", "has_opiates_dx")

map_dfr(zv_candidates, \(col) {
  x <- analysis_tibble[[col]]
  tibble(
    column     = col,
    n_distinct = n_distinct(x),
    pct_mode   = round(max(table(x, useNA = "no")) / length(x) * 100, 1)
  )
}) |> kable()
```

## Redundancy Checks

### withdrawal_pre vs withdrawal_pre_score

Both appear to be the same pre-induction withdrawal severity scalar. If identical,
keep `withdrawal_pre_score` (the engineered version from §21) and drop `withdrawal_pre`.

```{r redundancy-withdrawal}
cat("Identical?\n")
print(identical(analysis_tibble$withdrawal_pre, analysis_tibble$withdrawal_pre_score))
cat("\nCorrelation:\n")
print(cor(analysis_tibble$withdrawal_pre, analysis_tibble$withdrawal_pre_score,
          use = "complete.obs"))
cat("\nCrosstab (should be diagonal if identical):\n")
print(table(analysis_tibble$withdrawal_pre, analysis_tibble$withdrawal_pre_score,
            useNA = "always"))
```

### depression_binary / anxiety_binary / schizophrenia_binary vs has_* flags

These may be from different sources (e.g., ASI self-report vs SCID diagnosis).
If identical → one is redundant. If different → both add information.

```{r redundancy-psych}
cat("depression_binary identical to has_major_dep?\n")
print(identical(analysis_tibble$depression_binary, analysis_tibble$has_major_dep))

cat("\nanxiety_binary identical to has_anx_pan?\n")
print(identical(analysis_tibble$anxiety_binary, analysis_tibble$has_anx_pan))

cat("\nschizophrenia_binary identical to has_schizophrenia?\n")
print(identical(analysis_tibble$schizophrenia_binary, analysis_tibble$has_schizophrenia))

# If not identical, show disagreement counts
cat("\ndepression vs has_major_dep disagreements:",
    sum(analysis_tibble$depression_binary != analysis_tibble$has_major_dep, na.rm = TRUE), "\n")
cat("anxiety vs has_anx_pan disagreements:",
    sum(analysis_tibble$anxiety_binary != analysis_tibble$has_anx_pan, na.rm = TRUE), "\n")
cat("schizophrenia disagreements:",
    sum(analysis_tibble$schizophrenia_binary != analysis_tibble$has_schizophrenia, na.rm = TRUE), "\n")
```

### dep_or_anx_binary vs depression_binary | anxiety_binary

```{r redundancy-dep-or-anx}
derived <- as.integer(
  analysis_tibble$depression_binary == 1 | analysis_tibble$anxiety_binary == 1
)
cat("dep_or_anx_binary == (depression_binary | anxiety_binary)?\n")
print(identical(analysis_tibble$dep_or_anx_binary, derived))
```

### marital vs is_married

`is_married` is likely derived from `marital` (1 if "Married or Partnered"). If so,
one is redundant when both are included. Cross-tab confirms the mapping.

```{r redundancy-marital}
cat("Crosstab of marital vs is_married:\n")
print(table(analysis_tibble$marital, analysis_tibble$is_married, useNA = "always"))
```

## Factor Levels

How many levels, and what are they? Important for deciding how to dummy encode and
whether `site` (many levels) needs special treatment.

```{r factor-levels}
fct_cols <- c("project", "race", "job", "education", "marital", "treatment_arm", "site")

for (col in fct_cols) {
  cat("\n===", col, "===\n")
  cat("n levels:", nlevels(analysis_tibble[[col]]), "\n")
  print(table(analysis_tibble[[col]], useNA = "always"))
}
```

## Ordered Factor Levels and NAs

```{r ordered-levels}
ord_cols <- c("ftnd_score", "per_day", "housing_stability_ord")

for (col in ord_cols) {
  cat("\n===", col, "===\n")
  cat("Levels:", paste(levels(analysis_tibble[[col]]), collapse = " < "), "\n")
  cat("n_na:", sum(is.na(analysis_tibble[[col]])), "\n")
  print(table(analysis_tibble[[col]], useNA = "always"))
}
```

## NA Checks — Columns Not Showing NAs in Glimpse

Verify these are truly complete before assuming `no_treatment` with no imputation.

```{r na-checks}
check_cols <- c(
  "pain_score", "withdrawal_pre", "withdrawal_pre_score",
  "iv_days_total", "iv_events", "iv_amount_dominant",
  "cocaine_inject_days", "heroin_inject_days", "speedball_inject_days",
  "opioid_inject_days", "speed_inject_days",
  "rx_days", "rx_categories", "rx_any_binary",
  "took_own_study_drug", "took_other_study_drug"
)

tibble(
  column = check_cols,
  n_na   = map_int(check_cols, \(col) sum(is.na(analysis_tibble[[col]]))),
  min    = map_dbl(check_cols, \(col) min(analysis_tibble[[col]], na.rm = TRUE)),
  max    = map_dbl(check_cols, \(col) max(analysis_tibble[[col]], na.rm = TRUE))
) |> kable()
```

## alcohol_missing_amnt_* — What Does "Missing Amount" Mean?

In `all_drugs`, alcohol entries where the amount category (heavy/light) wasn't
recorded are grouped as "missing." This is a data artifact, not a meaningful
predictor in the usual sense. Check prevalence.

```{r alcohol-missing}
cat("alcohol_missing_amnt_days — range:",
    range(analysis_tibble$alcohol_missing_amnt_days), "\n")
cat("n participants with any missing-amount days:",
    sum(analysis_tibble$alcohol_missing_amnt_days > 0), "\n")
cat("alcohol_missing_amnt_binary — n=1:",
    sum(analysis_tibble$alcohol_missing_amnt_binary), "\n")
```

## iv_amount_dominant — Extreme Values

Values up to 1500 visible in glimpse. Check distribution shape. If heavily
right-skewed, a log1p transform might help logistic regression.

```{r iv-amount}
summary(analysis_tibble$iv_amount_dominant)
hist(analysis_tibble$iv_amount_dominant, breaks = 50,
     main = "iv_amount_dominant distribution", xlab = "amount")
```

---

# Per-Column Sections (Glimpse Order)

Groups used below:
- `ID` — exclude via `update_role()`
- `no_treatment` — numeric/int, no NAs, no encoding needed
- `nominal_factor` — unordered `<fct>`, one-hot dummy encode
- `ordered_factor` — `<ord>`, step_ordinalscore() or polynomial contrasts
- `mcar` — item-level missingness, kNN impute
- `structural_na` — study-level MNAR, step_indicate_na() + median impute
- `check_zv` — suspected near-zero variance, verify before including
- `check_redundant` — potentially duplicates another column

---

## who
**Group:** `ID` — exclude via `update_role(who, new_role = "ID")`.

## project
**Group:** `nominal_factor` — `<fct>`, 3 levels (27/30/51). One-hot encode. No NAs (all participants assigned to a study).

## alcohol_heavy_amnt_days
**Group:** `no_treatment` — `<int>`, count 0–28.

## alcohol_heavy_amnt_binary
**Group:** `no_treatment` — `<int>`, 0/1.

## alcohol_heavy_amnt_streak
**Group:** `no_treatment` — `<int>`, count.

## alcohol_light_amnt_days
**Group:** `no_treatment` — `<int>`, count 0–28.

## alcohol_light_amnt_binary
**Group:** `no_treatment` — `<int>`, 0/1.

## alcohol_missing_amnt_days
**Group:** `no_treatment` (tentative) — `<int>`. Days of alcohol use with uncategorized amount. See spot check above — if very few non-zero, may approach ZV.

## alcohol_missing_amnt_binary
**Group:** `no_treatment` (tentative) — `<int>`, 0/1. Same caveat.

## amphetamine_days
**Group:** `no_treatment` — `<int>`, count.

## amphetamine_binary
**Group:** `no_treatment` — `<int>`, 0/1.

## amphetamine_streak
**Group:** `no_treatment` — `<int>`, count.

## analgesic_days
**Group:** `no_treatment` — `<int>`, count. Low expected prevalence.

## analgesic_binary
**Group:** `no_treatment` — `<int>`, 0/1.

## antiemetic_days
**Group:** `no_treatment` — `<int>`, count. Low expected prevalence.

## antiemetic_binary
**Group:** `no_treatment` — `<int>`, 0/1.

## benzodiazepine_days
**Group:** `no_treatment` — `<int>`, count 0–28.

## benzodiazepine_binary
**Group:** `no_treatment` — `<int>`, 0/1.

## benzodiazepine_streak
**Group:** `no_treatment` — `<int>`, count.

## buprenorphine_days
**Group:** `no_treatment` — `<int>`, count. Pre-randomization BUP use.

## buprenorphine_binary
**Group:** `no_treatment` — `<int>`, 0/1.

## cannabinoids_days
**Group:** `no_treatment` — `<int>`, count.

## cannabinoids_binary
**Group:** `no_treatment` — `<int>`, 0/1.

## cannabinoids_streak
**Group:** `no_treatment` — `<int>`, count.

## cocaine_days
**Group:** `no_treatment` — `<int>`, count.

## cocaine_binary
**Group:** `no_treatment` — `<int>`, 0/1.

## cocaine_streak
**Group:** `no_treatment` — `<int>`, count.

## crack_days
**Group:** `no_treatment` — `<int>`, count.

## crack_binary
**Group:** `no_treatment` — `<int>`, 0/1.

## crack_streak
**Group:** `no_treatment` — `<int>`, count.

## fentanyl_days
**Group:** `no_treatment` (tentative, low prevalence) — `<int>`, count. Batch audit will show whether n_distinct = 1 (ZV).

## fentanyl_binary
**Group:** `no_treatment` (tentative) — `<int>`, 0/1.

## fentanyl_streak
**Group:** `no_treatment` (tentative) — `<int>`, count.

## heroin_days
**Group:** `check_zv` — `<int>`. Glimpse shows many 28s (max possible). If the vast majority of participants have 28, this approaches ZV. See ZV spot check.

## heroin_binary
**Group:** `check_zv` — `<int>`, 0/1. Glimpse shows all 1s — OUD inclusion criterion makes heroin use near-universal. If ZV, `step_zv()` removes it automatically.

## heroin_streak
**Group:** `no_treatment` — `<int>`, count. More variable than heroin_binary.

## mdma_hallucinogen_days
**Group:** `no_treatment` (low prevalence) — `<int>`, count.

## mdma_hallucinogen_binary
**Group:** `no_treatment` (low prevalence) — `<int>`, 0/1.

## methadone_days
**Group:** `no_treatment` — `<int>`, count. Pre-randomization methadone use.

## methadone_binary
**Group:** `no_treatment` — `<int>`, 0/1.

## methamphetamine_days
**Group:** `no_treatment` — `<int>`, count.

## methamphetamine_binary
**Group:** `no_treatment` — `<int>`, 0/1.

## methamphetamine_streak
**Group:** `no_treatment` — `<int>`, count.

## muscle_relaxant_days
**Group:** `no_treatment` (low prevalence) — `<int>`, count.

## muscle_relaxant_binary
**Group:** `no_treatment` (low prevalence) — `<int>`, 0/1.

## opioid_days
**Group:** `no_treatment` — `<int>`, count. Prescription opioids (distinct from heroin).

## opioid_binary
**Group:** `no_treatment` — `<int>`, 0/1.

## opioid_streak
**Group:** `no_treatment` — `<int>`, count.

## pcp_days
**Group:** `no_treatment` (very low prevalence) — `<int>`, count. Likely near-ZV.

## pcp_binary
**Group:** `no_treatment` (very low prevalence) — `<int>`, 0/1.

## pcp_streak
**Group:** `no_treatment` — `<int>`, count.

## sedatives_days
**Group:** `no_treatment` — `<int>`, count.

## sedatives_binary
**Group:** `no_treatment` — `<int>`, 0/1.

## sedatives_streak
**Group:** `no_treatment` — `<int>`, count.

## suboxone_days
**Group:** `no_treatment` — `<int>`, count. Pre-randomization suboxone.

## suboxone_binary
**Group:** `no_treatment` — `<int>`, 0/1.

## alcohol_restraint
**Group:** `no_treatment` — `<int>`, count (days of alcohol abstinence in the window).

## polydrug_days
**Group:** `no_treatment` — `<int>`, count.

## drug_breadth
**Group:** `no_treatment` — `<int>`, count of distinct drug categories used.

## asi_iv_binary
**Group:** `no_treatment` — `<int>`, 0/1. ASI IV drug use flag.

## age
**Group:** `no_treatment` — `<dbl>`, continuous. No NAs expected.

## is_hispanic
**Group:** `no_treatment` — `<int>`, 0/1.

## race
**Group:** `nominal_factor` — `<fct>`. Unordered. One-hot encode. See factor levels spot check for full level list.

## job
**Group:** `nominal_factor` + `mcar` — `<fct>`, has NAs. kNN impute first, then one-hot encode.

## is_living_stable
**Group:** `mcar` — `<int>`, 0/1, has NAs. kNN impute (or mode impute).

## education
**Group:** `nominal_factor` + `mcar` — `<fct>`, has NAs. kNN impute, then one-hot encode.

## marital
**Group:** `nominal_factor` + `mcar` + `check_redundant` — `<fct>`, has NAs. See redundancy check vs `is_married`. If keeping, kNN impute then one-hot encode.

## is_married
**Group:** `mcar` + `check_redundant` — `<int>`, 0/1, has NAs. Likely derived from `marital` → redundant if `marital` is in the recipe. See redundancy spot check.

## is_male
**Group:** `no_treatment` — `<int>`, 0/1. No NAs expected.

## is_smoker
**Group:** `check_zv` — `<int>`, 0/1. Glimpse shows all 1s — OUD populations have extremely high smoking rates. If ZV, `step_zv()` removes it. See ZV spot check.

## ftnd_score
**Group:** `ordered_factor` — `<ord>`. Fagerström nicotine dependence score. Linear spacing defensible (0–10 scale). Check NAs (see spot check). If NAs exist, add `mcar`. Use `step_ordinalscore()`.

## per_day
**Group:** `ordered_factor` — `<ord>`. Cigarettes per day in ordered categories ("10 OR LESS", "11-20", "21-30", "31+"). Spacing between categories is unequal — consider polynomial contrasts (`step_dummy(one_hot = FALSE)`) instead of `step_ordinalscore()`. Check NAs. See levels spot check.

## pain_score
**Group:** `no_treatment` (tentative) — `<int>`. Glimpse shows 0/1/2 range. Treated as continuous integer. See NA and range check above.

## has_schizophrenia
**Group:** `no_treatment` — `<int>`, 0/1.

## has_major_dep
**Group:** `no_treatment` + `check_redundant` — `<int>`, 0/1. See redundancy check vs `depression_binary`.

## has_bipolar
**Group:** `no_treatment` — `<int>`, 0/1.

## has_anx_pan
**Group:** `no_treatment` + `check_redundant` — `<int>`, 0/1. See redundancy check vs `anxiety_binary`.

## has_brain_damage
**Group:** `no_treatment` — `<int>`, 0/1. Low prevalence expected.

## has_epilepsy
**Group:** `no_treatment` — `<int>`, 0/1. Low prevalence expected.

## has_opiates_dx
**Group:** `check_zv` — `<int>`, 0/1. OUD diagnosis is inclusion criterion — likely all 1s. If ZV, exclude. See ZV spot check.

## has_alcol_dx
**Group:** `structural_na` — `<int>`, MNAR (study-level). `step_indicate_na()` + `step_impute_median()`.

## has_amphetamines_dx
**Group:** `structural_na` — `<int>`, MNAR.

## has_cannabis_dx
**Group:** `structural_na` — `<int>`, MNAR.

## has_cocaine_dx
**Group:** `structural_na` — `<int>`, MNAR.

## has_sedatives_dx
**Group:** `structural_na` — `<int>`, MNAR.

## depression_binary
**Group:** `check_redundant` — `<int>`, 0/1. May be identical to `has_major_dep`. See redundancy spot check. Final group depends on result: if identical → drop; if different → `no_treatment`.

## anxiety_binary
**Group:** `check_redundant` — `<int>`, 0/1. May be identical to `has_anx_pan`. See redundancy spot check.

## schizophrenia_binary
**Group:** `check_redundant` — `<int>`, 0/1. May be identical to `has_schizophrenia`.

## psych_comorbidity_count
**Group:** `no_treatment` — `<int>`, count. Derived from has_* flags. Correlated with them but not redundant (it's a sum).

## substance_dx_count
**Group:** `structural_na` — `<int>`, MNAR. Missing for same studies as `has_*_dx`.

## dep_or_anx_binary
**Group:** `check_redundant` — `<int>`, 0/1. Likely = `depression_binary | anxiety_binary`. If so, redundant with both. See redundancy spot check.

## is_homeless
**Group:** `structural_na` — `<int>`, MNAR. 77% NA (study-level, not item-level). 570 observed values worth keeping. `step_indicate_na()` + `step_impute_median()`.

## housing_stability_ord
**Group:** `ordered_factor` + `mcar` — `<ord>`, has NAs. kNN impute first, then `step_ordinalscore()` (or polynomial contrasts depending on levels). See ordered levels spot check.

## treatment_arm
**Group:** `nominal_factor` — `<fct>`. No NAs expected. One-hot encode. See factor levels spot check.

## heroin_rbs_alldr_incons
**Group:** `no_treatment` — `<int>`, 0/1. Heroin denial flag (TLFB × all-drugs inconsistency).

## iv_days_total
**Group:** `no_treatment` (pending NA check) — `<dbl>`. See NA spot check.

## iv_events
**Group:** `no_treatment` (pending NA check) — `<dbl>`. See NA spot check.

## iv_amount_dominant
**Group:** `no_treatment` (pending NA check and distribution check) — `<dbl>`. Extreme values visible. See iv_amount spot check. May benefit from log1p transform if right-skewed.

## iv_shared_needles
**Group:** `no_treatment` — `<int>`, 0/1.

## cocaine_inject_days
**Group:** `no_treatment` (pending NA check) — `<dbl>`. See NA spot check.

## heroin_inject_days
**Group:** `no_treatment` (pending NA check) — `<dbl>`. See NA spot check.

## speedball_inject_days
**Group:** `no_treatment` (pending NA check) — `<dbl>`. See NA spot check.

## opioid_inject_days
**Group:** `no_treatment` (pending NA check) — `<dbl>`. See NA spot check.

## speed_inject_days
**Group:** `no_treatment` (pending NA check) — `<dbl>`. See NA spot check.

## site
**Group:** `nominal_factor` — `<fct>`. Many clinical site levels. See factor levels spot check for exact count. High level count means many dummy columns — may want to consider whether to include.

## withdrawal_pre
**Group:** `check_redundant` — `<int>`. Appears identical to `withdrawal_pre_score`. See redundancy spot check. If identical → drop `withdrawal_pre`, keep `withdrawal_pre_score` (engineered version).

## took_own_study_drug
**Group:** `no_treatment` — `<int>`, 0/1. Whether participant took their own assigned study drug.

## took_other_study_drug
**Group:** `no_treatment` — `<int>`, 0/1.

## withdrawal_pre_score
**Group:** `no_treatment` (pending NA check) — `<int>`, 0–3 scalar. The engineered scalar from §21. See NA spot check and redundancy check vs `withdrawal_pre`.

## has_major_dep_x_withdrawal_pre_score
**Group:** `no_treatment` — `<int>`. Interaction term (already computed). Treat as numeric predictor.

## has_anx_pan_x_withdrawal_pre_score
**Group:** `no_treatment` — `<int>`. Interaction term.

## has_anx_pan_x_benzodiazepine_days
**Group:** `no_treatment` — `<int>`. Interaction term.

## rx_days
**Group:** `no_treatment` (pending NA check) — `<int>`. Prescription drug days in window. See NA spot check.

## rx_categories
**Group:** `no_treatment` (pending NA check) — `<int>`. Count of distinct prescription categories. See NA spot check.

## rx_any_binary
**Group:** `no_treatment` (pending NA check) — `<int>`, 0/1.
